---
title: "PBMC cell type annotation"
output: html_notebook
---

This notebook is a follow-up to 1_pbmc_preprocessing.Rmd. After preprocessing our PBMC data, we are ready to perform multimodal clustering and cell type annotation.

First, clear the environment and set the path where the project folder is saved.

```{r}
# Clear all objects from environment
rm(list = ls(all.names = TRUE))
# Free up memory
gc()
# Increase max vector memory
mem.maxVSize(200000)

project_path <- "/Users/juliaelrod/Desktop/cmu/research/scleroderma/jSSc_ASCT_monocytes_manuscript/code_for_manuscript/"
```

Now load in useful packages and custom functions.

```{r}
# Load packages
packages <- c("Seurat", "dplyr", "tidyr", "SeuratDisk", "harmony", "readr",
              "ggplot2", "stringr", "SoupX", "dsb", "data.table", "Matrix",
              "scDblFinder", "scater", "RColorBrewer")
invisible(lapply(packages, require, character.only = TRUE))

# Source functions
function_files <- list.files(paste0(project_path, "functions"), pattern = "\\.R$", full.names = TRUE)
invisible(lapply(function_files, source))

# Function to update cell_types
update_cell_types <- function(df, cluster, cell_type = NA) {
  df[df$seurat_clusters == cluster, "cell_type"] <- cell_type
  return(df)
}
```

Now, we load in the Seurat object we generated in the previous notebook.

```{r}
# # Load Seurat object
# scler_pbmc <- LoadH5Seurat(paste0(project_path, "data/h5seurat/scler_pbmc_8.h5seurat"))
# 
# # RNA harmony reduction is ready to be used with WNN.
# 
# # Now, set up integrated DSB values to use in WNN analysis.
# DefaultAssay(scler_pbmc) <- "DSB_integrated"
# 
# # Scale and run PCA on integrated DSB data
# scler_pbmc <- scler_pbmc %>%
#   ScaleData() %>%
#   RunPCA(reduction.name = 'pca_dsb_int', verbose = FALSE)
# 
# # Run WSNN
# scler_pbmc <- FindMultiModalNeighbors(
#   scler_pbmc, reduction.list = list("harmony", "pca_dsb_int"),
#   dims.list = list(1:30, 1:18),
#   verbose = FALSE,
#   k.nn = 20
# )
# 
# # Run UMAP
# scler_pbmc <- RunUMAP(scler_pbmc, 
#                       nn.name = "weighted.nn", 
#                       reduction.name = "wnn.umap", 
#                       reduction.key = "wnnUMAP_")
# 
# 
# # Cluster using SLM algorithm (3)
# scler_pbmc <- FindClusters(scler_pbmc, graph.name = "wsnn",
#                   algorithm = 3,
#                   resolution = 0.3,
#                   random.seed = 1990)
# 
# # Save as H5
# SaveH5Seurat(scler_pbmc, paste0(project_path, "data/h5seurat/scler_pbmc_9"))
```

Now, we use Seurat's FindMarkers function to get differentially expressed genes and proteins for each cluster.

```{r}
# # # Load data
# scler_pbmc <- LoadH5Seurat(paste0(project_path, "data/h5seurat/scler_pbmc_9.h5seurat"))
# # 
# # # Take a look
# DimPlot(scler_pbmc, label = T, repel = T)
# # 
# # # Get markers by cluster
# # 
# # # # RNA markers
# # # rna_de <- FindAllMarkers(scler_pbmc, assay = "RNA", only.pos = T)
# # # # Save as RDS
# # # saveRDS(rna_de, paste0(project_path, "data/RDS_files/markers/rna_de.RDS"))
# # # Read RDS
# rna_de <- readRDS(paste0(project_path, "data/RDS_files/markers/rna_de.RDS"))
# # 
# # # # ADT markers
# # # dsb_de <- FindAllMarkers(scler_pbmc, assay = "DSB", only.pos = T)
# # # # Save as RDS
# # # saveRDS(dsb_de, paste0(project_path, "data/RDS_files/markers/dsb_de.RDS"))
# # # Read RDS
# dsb_de <- readRDS(paste0(project_path, "data/RDS_files/markers/dsb_de.RDS"))
```

Now we annotate each cluster based on canonical RNA and protein markers. 

```{r}
# # Data frame to store cell types
# cell_types <- data.frame(seurat_clusters = levels(scler_pbmc$seurat_clusters),
#                          cell_type = NA)
# ################################################################################
# #  CLUSTER 0                                                                   #
# #  TYPE: MONOCYTE                                                        #
# #  RNA MARKERS: CD14, LYZ                                                      #
# #  ADT MARKERS:                                                                #
# ################################################################################
# # Look at top markers for each layer
# display_top(rna_de, 0); display_top(dsb_de, 0)
# # Look at CD14
# rna_de %>% filter(gene == "CD14" & cluster == 0)
# # Update cell_types
# cell_types <- update_cell_types(cell_types, 0, cell_type = "monocyte")
# ################################################################################
# #  CLUSTER 1                                                                   #
# #  TYPE: B CELL                                                                #
# #  RNA MARKERS: CD79A, MS4A1, CD79B, IGHM                                      #
# #  ADT MARKERS: IgM                                                            #
# ################################################################################
# # Look at top markers for each layer
# display_top(rna_de, 1); display_top(dsb_de, 1)
# # Update cell_types
# cell_types <- update_cell_types(cell_types, 1, cell_type = "B cell")
# ################################################################################
# #  CLUSTER 2                                                                   #
# #  TYPE: CD4+ T CELL                                                           #
# #  RNA MARKERS: CD3E, CD3D, CD3G, CD4                                          #
# #  ADT MARKERS: CD2, CD7                                                       #
# ################################################################################
# # Look at top markers for each layer
# display_top(rna_de, 2); display_top(dsb_de, 2)
# # Look at CD4
# rna_de %>% filter(cluster == 2 & gene == "CD4")
# # Update cell_types
# cell_types <- update_cell_types(cell_types, 2, cell_type = "CD4+ T cell")
# ################################################################################
# #  CLUSTER 3                                                                   #
# #  TYPE: MONOCYTE                                                        #
# #  RNA MARKERS: FCGR3A                                                         #
# #  ADT MARKERS: CD16                                                           #
# ################################################################################
# # Look at top markers for each layer
# display_top(rna_de, 3); display_top(dsb_de, 3)
# # Look at FCGR3A (CD16)
# rna_de %>% filter(cluster == 3 & gene == "FCGR3A")
# # Update cell_types
# cell_types <- update_cell_types(cell_types, 3, cell_type = "monocyte")
# ################################################################################
# #  CLUSTER 4                                                                   #
# #  TYPE: NATURAL KILLER                                                        #
# #  RNA MARKERS: GNLY, NKG7, GZMA, GZMB, KLRB1, KLRG1                           #
# #  ADT MARKERS: CD56, CD16                                                     #
# ################################################################################
# # Look at top markers for each layer
# display_top(rna_de, 4); display_top(dsb_de, 4)
# # Update cell_types
# cell_types <- update_cell_types(cell_types, 4, cell_type = "natural killer")
# ################################################################################
# #  CLUSTER 5                                                                   #
# #  TYPE: CD8+ T CELL                                                           #
# #  RNA MARKERS: CD3E, CD3D, CD8B, CD8A, CD7                                    #
# #  ADT MARKERS: CD2, CD8, CD3, CD5, CD7                                        #
# ################################################################################
# # Look at top markers for each layer
# display_top(rna_de, 5); display_top(dsb_de, 5)
# # Update cell_types
# cell_types <- update_cell_types(cell_types, 5, cell_type = "CD8+ T cell")
# ################################################################################
# #  CLUSTER 6                                                                   #
# #  TYPE: NATURAL KILLER                                                        #
# #  RNA MARKERS: GNLY, KLRB1, NKG7, GZMA, GZMB, NCAM1                           #
# #  ADT MARKERS: CD56                                                           #
# ################################################################################
# # Look at top markers for each layer
# display_top(rna_de, 6); display_top(dsb_de, 6)
# # Look at NCAM1
# rna_de %>% filter(cluster == 6 & gene == "NCAM1")
# # Look at CD56
# dsb_de %>% filter(cluster == 6 & gene == "anti-human-CD56")
# # Update cell_types
# cell_types <- update_cell_types(cell_types, 6, cell_type = "natural killer")
# ################################################################################
# #  CLUSTER 7                                                                   #
# #  TYPE: MONOCYTE                                                        #
# #  RNA MARKERS: LYZ, CD14                                                      #
# #  ADT MARKERS:                                                                #
# ################################################################################
# # Look at top markers for each layer
# display_top(rna_de, 7); display_top(dsb_de, 7)
# # Update cell_types
# cell_types <- update_cell_types(cell_types, 7, cell_type = "monocyte")
# ################################################################################
# #  CLUSTER 8                                                                   #
# #  TYPE: B CELL                                                                #
# #  RNA MARKERS: CD79A, IGHM, MS4A1, IGHD                                       #
# #  ADT MARKERS: CD19, IgM, CD20, CD21, IgD                                     #
# ################################################################################
# # Look at top markers for each layer
# display_top(rna_de, 8); display_top(dsb_de, 8)
# # Update cell_types
# cell_types <- update_cell_types(cell_types, 8, cell_type = "B cell")
# ################################################################################
# #  CLUSTER 9                                                                   #
# #  TYPE: mDC                                                                   #
# #  RNA MARKERS: CD1C                                                           #
# #  ADT MARKERS: CD1C                                                           #
# ################################################################################
# # Look at top markers for each layer
# display_top(rna_de, 9); display_top(dsb_de, 9)
# # Look at CD1C
# rna_de %>% filter(cluster == 9 & gene == "CD1C"); dsb_de %>% filter(cluster == 9 & gene == "anti-human-CD1c")
# # Update cell_types
# cell_types <- update_cell_types(cell_types, 9, cell_type = "mDC")
# ################################################################################
# #  CLUSTER 10                                                                  #
# #  TYPE:  T/B BIPHENOTYPIC                                                     #
# #  RNA MARKERS: CD7, CD3E, CD79A, IGHM, MS4A1                                  #
# #  ADT MARKERS: CD19, CD21, IgD, CD20, IgM, CD2, CD5, CD3                      #
# ################################################################################
# # Look at top markers for each layer
# display_top(rna_de, 10); display_top(dsb_de, 10)
# # Update cell_types
# cell_types <- update_cell_types(cell_types, 10, cell_type = "T/B biphenotypic")
# #
# # Since this is an unusual cell type, let's see whether these are mainly patient cells
# clus10_counts_by_group <- scler_pbmc@meta.data %>% filter(seurat_clusters == 10) %>% pull(group) %>% table()
# all_counts_by_group <- scler_pbmc@meta.data %>% pull(group) %>% table()
# clus10_prop_by_group <- clus10_counts_by_group / all_counts_by_group
# clus10_prop_by_group
# # Yes, this cluster includes a greater proportion of patient cells at all time
# # points than healthy cells.
# ################################################################################
# #  CLUSTER 11                                                                  #
# #  Source for pDC markers:                                                     #
# #  https://www.beckman.com/resources/cell-types/blood-cells/leukocytes/        #
# #  dendritic-cells/plasmacytoid-cells                                          #
# #  TYPE: pDC                                                                   #
# #  RNA MARKERS: IL3RA (CD123)                                                  #
# #  ADT MARKERS: CD123                                                          #
# ################################################################################
# # Look at top markers for each layer
# display_top(rna_de, 11); display_top(dsb_de, 11)
# # Look at IL3RA
# rna_de %>% filter(cluster == 11 & gene == "IL3RA")
# # Update cell_types
# cell_types <- update_cell_types(cell_types, 11, cell_type = "pDC")
# ################################################################################
# #  CLUSTER 12                                                                  #
# #  TYPE: NATURAL KILLER/T CELL                                                 #
# #  RNA MARKERS: NCAM1                                                          #
# #  ADT MARKERS: CD7, CD2, CD3, CD5                                             #
# ################################################################################
# # Look at top markers for each layer
# display_top(rna_de, 12); display_top(dsb_de, 12)
# # Update cell_types
# cell_types <- update_cell_types(cell_types, 12, cell_type = "natural killer/T cell")
# ################################################################################
# #  CLUSTER 13                                                                  #
# #  TYPE: HSPC                                                                  #
# #  RNA MARKERS: CD34, MYB                                                      #
# #  ADT MARKERS:                                                                #
# ################################################################################
# # Look at top markers for each layer
# display_top(rna_de, 13); display_top(dsb_de, 13)
# # CD34 and MYB are HSPC markers
# hspc_genes <- c("CD34", "MYB")
# # Violin plot
# VlnPlot(scler_pbmc, hspc_genes)
# # Update cell_types
# cell_types <- update_cell_types(cell_types, 13, cell_type = "HSPC")
# ################################################################################
# #  CLUSTER 14                                                                  #
# #  TYPE: PLATELET                                                              #
# #  RNA MARKERS: PF4, PPBP, TUBB1, GP1BA, THBS1                                 #
# #  ADT MARKERS: CD41                                                           #
# ################################################################################
# # Look at top markers for each layer
# display_top(rna_de, 14); display_top(dsb_de, 14)
# # Define platelet markers
# platelet_genes <- c("PF4", "PPBP", "TUBB1", "GP1BA", "THBS1")
# # Visualize with violin plot
# VlnPlot(scler_pbmc, features = platelet_genes)
# # Update cell_types
# cell_types <- update_cell_types(cell_types, 14, cell_type = "platelet")
# ################################################################################
# #  CLUSTER 15                                                                  #
# #  TYPE: NATURAL KILLER                                                        #
# #  RNA MARKERS: NCAM1                                                          #
# #  ADT MARKERS: CD56, CD16                                                     #
# ################################################################################
# # Look at top markers for each layer
# display_top(rna_de, 15); display_top(dsb_de, 15)
# # Look at NCAM1
# rna_de %>% filter(cluster == 15 & gene == "NCAM1")
# # Update cell_types
# cell_types <- update_cell_types(cell_types, 15, cell_type = "natural killer")
# 
# # Add cell types to Seurat object
# cell_type_vec <-
#    setNames(as.character(cell_types$cell_type),
#           as.character(cell_types$seurat_clusters))[as.character(scler_pbmc$seurat_clusters)]
# names(cell_type_vec) <- NULL
# scler_pbmc$cell_type <- cell_type_vec
# 
# # Plot it
# s <- DimPlot(scler_pbmc,
#         reduction = "wnn.umap",
#         group.by = "cell_type",
#         pt.size = 2,
#         alpha = 0.8,
#         cols = my_colors,
#         combine = F)[[1]]
# # Adjust labels
# t <- s + labs(title = "jSSc and healthy control annotated PBMC clusters") + xlab("UMAP coordinate 1") + ylab("UMAP coordinate 2")
# # Take a look
# t
# 
# # Look at cell type proportions:
# 
# # Total cells
# total_cells <- ncol(scler_pbmc)
# 
# # Number of cells per type
# num_cells_per_type <- table(scler_pbmc$cell_type)
# 
# # Pct of cells per type
# pct_cells_per_type <- as.matrix(paste0((round((num_cells_per_type / total_cells), 3) * 100), " %"))
# rownames(pct_cells_per_type) <- names(num_cells_per_type)
# pct_cells_per_type
# 
# # Save as PNG
# umap_path <- paste0(project_path, "figures/")
# ggsave(paste0(umap_path, "pbmc_umap", Sys.Date(), ".png"), t, height = 7, width = 9, bg = "white")
# 
# # Save as H5
# SaveH5Seurat(scler_pbmc, paste0(project_path, "data/h5seurat/scler_pbmc_10"))
```

UMAP

```{r}
scler_pbmc <- LoadH5Seurat(paste0(project_path, "data/h5seurat/scler_pbmc_10.h5seurat"))

Idents(scler_pbmc) <- scler_pbmc$cell_type
all_umap <- 
scler_pbmc %>% DimPlot(label = T, cols = brewer.pal(12, "Set3"), repel = T) + ggtitle("jSSc and healthy control annotated PBMC clusters")
ggsave(paste0(project_path, "figures/all_umap.png"), all_umap, height = 6, width = 7)

all_umap_by_health <-
scler_pbmc %>% DimPlot(label = T, cols = brewer.pal(12, "Set3"), repel = T, split.by = "health") + ggtitle("Healthy control vs. jSSc annotated PBMC clusters")
ggsave(paste0(project_path, "figures/all_umap_by_health.png"), all_umap_by_health, width = 9, height = 4.5)

all_umap_by_group <-
scler_pbmc %>% DimPlot(label = T, cols = brewer.pal(12, "Set3"), repel = T, split.by = "group") + ggtitle("Healthy control vs. jSSc over time since ASCT annotated PBMC clusters")
ggsave(paste0(project_path, "figures/all_umap_by_group.png"), all_umap_by_group, width = 18, height = 4.5)

scler_pbmc$LIBRARY <- factor(scler_pbmc$LIBRARY, levels = c("H", "515", "591", "594"))
levels(scler_pbmc$LIBRARY)[levels(scler_pbmc$LIBRARY) == "H"] <- "HC"
all_umap_by_library <-
scler_pbmc %>% DimPlot(label = T, cols = brewer.pal(12, "Set3"), repel = T, split.by = "LIBRARY") + ggtitle("Healthy control vs. jSSc patient PBMC clusters")
ggsave(paste0(project_path, "figures/all_umap_by_library.png"), all_umap_by_library, width = 10, height = 5.72)

scler_pbmc$cell_type <- factor(scler_pbmc$cell_type,
                               levels = c("monocyte",
                                          "B cell",
                                          "T/B biphenotypic",
                                          "CD4+ T cell",
                                          "CD8+ T cell",
                                          "natural killer/T cell",
                                          "natural killer",
                                          "mDC",
                                          "pDC",
                                          "HSPC",
                                          "platelet"))

# Add a blank level between HC and SSc baseline
scler_pbmc@meta.data$group_spaced <- factor(
  scler_pbmc@meta.data$group,
  levels = c(levels(scler_pbmc@meta.data$group)[1], "", 
             levels(scler_pbmc@meta.data$group)[2:5])
)

# Plot it
prop_by_group <- ggplot(scler_pbmc@meta.data, aes(y = group_spaced, fill = cell_type)) +
    geom_bar(position = "fill", color = "black") +
    scale_fill_brewer(palette = "Set3", name = "Cell type") +
    scale_y_discrete(labels = c("HC", "", "SSc baseline", "SSc 6 mo.", "SSc 12 mo.", "SSc 24 mo."),
                     drop = FALSE) +
    coord_flip() +
    theme_bw() +
    labs(x = "Cell type proportion", y = "Group") +
    geom_hline(yintercept = 2, col = "black", linewidth = 1, linetype = "dashed") +
    ggtitle("HC vs. jSSc cell type proportions over time since ASCT") +
    theme(
        axis.text.x = element_text(size = 16, angle = 45, hjust = 1, color = "black"),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_text(size = 18),
        axis.title.y = element_text(size = 18),
        plot.title = element_text(size = 18, face = "bold"),
        legend.text = element_text(size = 14)
    )
ggsave(paste0(project_path, "figures/prop_by_group.png"), prop_by_group, height = 7, width = 8)

```



Heatmap showing the markers we used above

```{r}
# Load Seurat objects with labeled cell types
scler_pbmc <- LoadH5Seurat(paste0(project_path, "data/h5seurat/scler_pbmc_10.h5seurat"))

# FindAllMarkers by cell type
Idents(scler_pbmc) <- "cell_type"
cell_type_markers_RNA <- FindAllMarkers(scler_pbmc, assay = "RNA")
cell_type_markers_ADT <- FindAllMarkers(scler_pbmc, assay = "DSB")

# Monocyte markers
# Lafyatis paper cites FCN1, S100A8, and S100A9 as monocyte markers
# https://acrjournals.onlinelibrary.wiley.com/doi/full/10.1002/art.41813?saml_referrer 
monocyte_markers_RNA <- c("FCN1", "S100A8", "S100A9")
monocyte_markers_ADT <- c("anti-human-CD14")

# B cell markers
# CD22: https://www.biocompare.com/Editorial-Articles/570578-A-Guide-to-B-Cell-Markers/
B_cell_markers_RNA <- c("CD79A", "MS4A1", "BANK1")
B_cell_markers_ADT <- c("anti-human-CD19")

# T cell markers
T_cell_markers_RNA <- c("CD3D", "CD3E")
CD4_T_cell_markers_RNA <- c("IL7R")
CD8_T_cell_markers_RNA <- c("CD8A", "CD8B")
T_cell_markers_ADT <- c("anti-human-CD3")
CD4_T_cell_markers_ADT <- c("anti-human-CD4")
CD8_T_cell_markers_ADT <- c("anti-human-CD8")

# natural killer/T markers
NK_T_markers_RNA <- c("STAT4", "RORA")

# NK markers
NK_markers_RNA <- c("GNLY", "NKG7", "KLRB1")
NK_markers_ADT <- c("anti-human-CD56", "anti-human-CD16")

# mDC markers
# https://www.biocompare.com/Editorial-Articles/572982-Dendritic-Cell-Markers/
mDC_markers_RNA <- c("CD1C", "CLEC10A")
mDC_markers_ADT <- c("anti-human-HLA-DR")

# pDC markers
# https://www.biocompare.com/Editorial-Articles/572982-Dendritic-Cell-Markers/
# https://en.wikipedia.org/wiki/LILRA4
pDC_markers_RNA <- c("IL3RA", "LILRA4", "CLEC4C")
pDC_markers_ADT <- c("anti-human-CD123")

# HSPC markers
# https://pmc.ncbi.nlm.nih.gov/articles/PMC9564379/
HSPC_markers_RNA <- c("CD34", "MYB", "KIT")
HSPC_markers_ADT <- c("anti-human-CD99")

# Platelet markers
platelet_markers_RNA <- c("PF4", "PPBP", "TUBB1")
platelet_markers_ADT <- c("anti-human-CD41")

# Put all RNA markers together
rna_markers <- c(monocyte_markers_RNA,
                 B_cell_markers_RNA,
                 T_cell_markers_RNA,
                 CD4_T_cell_markers_RNA,
                 CD8_T_cell_markers_RNA,
                 NK_T_markers_RNA,
                 NK_markers_RNA,
                 mDC_markers_RNA,
                 pDC_markers_RNA,
                 HSPC_markers_RNA,
                 platelet_markers_RNA)

# Put all ADT markers together
adt_markers <- c(monocyte_markers_ADT,
                 B_cell_markers_ADT,
                 T_cell_markers_ADT,
                 CD4_T_cell_markers_ADT,
                 CD8_T_cell_markers_ADT,
                 NK_markers_ADT,
                 mDC_markers_ADT,
                 pDC_markers_ADT,
                 HSPC_markers_ADT,
                 platelet_markers_ADT)

# Let's put our cell types in the same order as the markers
scler_pbmc$cell_type <- factor(scler_pbmc$cell_type,
                               levels = c("monocyte",
                                          "B cell",
                                          "T/B biphenotypic",
                                          "CD4+ T cell",
                                          "CD8+ T cell",
                                          "natural killer/T cell",
                                          "natural killer",
                                          "mDC",
                                          "pDC",
                                          "HSPC",
                                          "platelet"))


Idents(scler_pbmc) <- scler_pbmc$cell_type

rna_marker_heatmap <- 
    DoHeatmap(subset(scler_pbmc, downsample = 100),
              features = rna_markers,
              assay = "RNA",
              group.bar.height = 0,
              size = 5,
              lines.width = 3) +
    scale_fill_gradient(low = "grey95", high = "darkblue", na.value = "white") +
  ggtitle("RNA markers") +
    theme(legend.position = "right",
          axis.text.y = element_text(size = 15, color = "black"),
          plot.title = element_text(size = 25, face = "bold")) +
    guides(color = "none")
ggsave(paste0(project_path, "figures/rna_marker_heatmap.png"), rna_marker_heatmap, width = 7, height = 9)

label_map <- setNames(gsub("^anti-human-", "", adt_markers), adt_markers)
adt_marker_heatmap <- 
    DoHeatmap(subset(scler_pbmc, downsample = 100),
              features = adt_markers,
              assay = "DSB",
              group.bar.height = 0,
              size = 5,
              lines.width = 8,
              disp.min = -0.5,
              disp.max = 1) +
    scale_fill_gradient(low = "grey95", high = "darkred", na.value = "black") +
    scale_y_discrete(labels = label_map) +
  ggtitle("ADT markers") +
    theme(legend.position = "right",
          axis.text.y = element_text(size = 15, color = "black"),
          plot.title = element_text(size = 25, face = "bold")) +
    guides(color = "none")
ggsave(paste0(project_path, "figures/adt_marker_heatmap.png"), adt_marker_heatmap, width = 7, height = 9)

rna_marker_dotplot <- 
  DotPlot(scler_pbmc,
          features = rna_markers,
          group.by = "cell_type",
          assay = "RNA") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
ggsave(paste0(project_path, "figures/rna_marker_dotplot.png"), rna_marker_dotplot, width = 12, height = 8)

adt_marker_dotplot <- 
  DotPlot(scler_pbmc,
          features = adt_markers,
          group.by = "cell_type",
          assay = "DSB",
          cols = c("lightgrey", "darkred")) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
ggsave(paste0(project_path, "figures/adt_marker_dotplot.png"), adt_marker_dotplot, width = 12, height = 8)


```


Now, let's try clustering the HC separately, and each patient individually.

```{r}
# # Load Seurat object
# scler_pbmc <- LoadH5Seurat(paste0(project_path, "data/h5seurat/scler_pbmc_8.h5seurat"))
# 
# # Default assay for clustering
# DefaultAssay(scler_pbmc) <- "DSB_integrated"
# 
# # Now, we create separate objects for the HC, and each individual patient
# healthy_pbmc <- scler_pbmc %>% subset(health == "healthy")
# p515_pbmc <- scler_pbmc %>% subset(subject == "515")
# p591_pbmc <- scler_pbmc %>% subset(subject == "591")
# p594_pbmc <- scler_pbmc %>% subset(subject == "594")
# 
# # Store in a list
# pbmc_list <- list(healthy_pbmc = healthy_pbmc,
#                   p515_pbmc = p515_pbmc,
#                   p591_pbmc = p591_pbmc,
#                   p594_pbmc = p594_pbmc)
# 
# # Scale and run PCA on integrated DSB data
# pbmc_list <- pbmc_list %>%
#   lapply(function(x) {
#     x %>%
#       ScaleData() %>%
#       RunPCA(reduction.name = "pca_dsb_int", verbose = FALSE) %>%
#       FindMultiModalNeighbors(reduction.list = list("harmony", "pca_dsb_int"),
#                               dims.list = list(1:30, 1:18),
#                               verbose = FALSE,
#                               k.nn = 20) %>%
#       RunUMAP(nn.name = "weighted.nn",
#               reduction.name = "wnn.umap",
#               reduction.key = "wnnUMAP_") %>%
#       FindClusters(graph.name = "wsnn",
#                   algorithm = 3,
#                   resolution = 0.3,
#                   random.seed = 1990)
#   })
# 
# 
# # Save as h5seurat
# lapply(names(pbmc_list), function(x){SaveH5Seurat(pbmc_list[[x]], paste0(project_path, "data/h5seurat/", x))})

# Read in
pbmc_list <- lapply(c("healthy_pbmc",
                      "p515_pbmc",
                      "p591_pbmc",
                      "p594_pbmc"),
                    function(x){
                      LoadH5Seurat(paste0(project_path, "data/h5seurat/", x, ".h5seurat"))
                    }) %>% setNames(c("healthy_pbmc",
                      "p515_pbmc",
                      "p591_pbmc",
                      "p594_pbmc"))

# Data frame to store cell types
healthy_cell_types <-data.frame(seurat_clusters = levels(pbmc_list$healthy_pbmc$seurat_clusters),
                         cell_type = NA)

# Take a look at each group
pbmc_list %>% lapply(function(x){DimPlot(x, label = T)})

# Find markers for each group
rna_marker_list <- pbmc_list %>% lapply(function(x){FindAllMarkers(x, assay = "RNA")})
adt_marker_list <- pbmc_list %>% lapply(function(x){FindAllMarkers(x, assay = "DSB")})

# Take a look at healthy markers

# 0
rna_marker_list$healthy_pbmc %>% display_top(0)
rna_marker_list$healthy_pbmc %>% filter(gene == "CD14")
healthy_cell_types <- update_cell_types(healthy_cell_types, 0, cell_type = "monocyte")

# 1
rna_marker_list$healthy_pbmc %>% display_top(1)
healthy_cell_types <- update_cell_types(healthy_cell_types, 1, cell_type = "B cell")

# 2
rna_marker_list$healthy_pbmc %>% display_top(2)
healthy_cell_types <- update_cell_types(healthy_cell_types, 2, cell_type = "T cell")

# 3
rna_marker_list$healthy_pbmc %>% display_top(3)
healthy_cell_types <- update_cell_types(healthy_cell_types, 3, cell_type = "monocyte")

# 4
rna_marker_list$healthy_pbmc %>% display_top(4)
adt_marker_list$healthy_pbmc %>% display_top(4)
healthy_cell_types <- update_cell_types(healthy_cell_types, 4, cell_type = "B cell")

# 5
rna_marker_list$healthy_pbmc %>% display_top(5)
adt_marker_list$healthy_pbmc %>% display_top(5)
healthy_cell_types <- update_cell_types(healthy_cell_types, 5, cell_type = "natural killer")

# 6
rna_marker_list$healthy_pbmc %>% display_top(6)
adt_marker_list$healthy_pbmc %>% display_top(6)
healthy_cell_types <- update_cell_types(healthy_cell_types, 6, cell_type = "monocyte")

# 7
rna_marker_list$healthy_pbmc %>% display_top(7)
adt_marker_list$healthy_pbmc %>% display_top(7)
healthy_cell_types <- update_cell_types(healthy_cell_types, 7, cell_type = "mDC")

# Add cell types to Seurat object
healthy_cell_type_vec <-
   setNames(as.character(healthy_cell_types$cell_type),
          as.character(healthy_cell_types$seurat_clusters))[as.character(pbmc_list$healthy_pbmc$seurat_clusters)]
names(healthy_cell_type_vec) <- NULL
pbmc_list$healthy_pbmc$cell_type <- healthy_cell_type_vec

# colors
n_col <- healthy_cell_type_vec %>% unique() %>% length()
COLORS <- my_colors[1:n_col]
names(COLORS) <- healthy_cell_type_vec %>% unique()

# Plot it
s <- DimPlot(pbmc_list$healthy_pbmc,
        reduction = "wnn.umap",
        group.by = "cell_type",
        pt.size = 2,
        alpha = 0.8,
        cols = COLORS, 
        combine = F)[[1]]
# Adjust labels
t <- s + labs(title = "Healthy control annotated PBMC clusters") + xlab("UMAP coordinate 1") + ylab("UMAP coordinate 2")
# Take a look
t

# Save as PNG
umap_path <- paste0(project_path, "figures/")
# ggsave(paste0(umap_path, "HEALTH_pbmc_umap", Sys.Date(), ".png"), t, height = 7, width = 9, bg = "white")

# Data frame to store cell types
p515_cell_types <-data.frame(seurat_clusters = levels(pbmc_list$p515_pbmc$seurat_clusters),
                         cell_type = NA)

# Take a look at 515 markers
# 0
rna_marker_list$p515_pbmc %>% display_top(0)
rna_marker_list$p515_pbmc %>% filter(gene == "CD14")
p515_cell_types <- update_cell_types(p515_cell_types, 0, cell_type = "monocyte")

# 1
rna_marker_list$p515_pbmc %>% display_top(1)
p515_cell_types <- update_cell_types(p515_cell_types, 1, cell_type = "B cell")

# 2
rna_marker_list$p515_pbmc %>% display_top(2)
p515_cell_types <- update_cell_types(p515_cell_types, 2, cell_type = "T cell")

# 3
rna_marker_list$p515_pbmc %>% display_top(3)
adt_marker_list$p515_pbmc %>% display_top(3)
rna_marker_list$p515_pbmc %>% filter(cluster == 3 & gene == "NCAM1")
p515_cell_types <- update_cell_types(p515_cell_types, 3, cell_type = "natural killer")

# 4
rna_marker_list$p515_pbmc %>% display_top(4)
p515_cell_types <- update_cell_types(p515_cell_types, 4, cell_type = "T cell")

# 5
rna_marker_list$p515_pbmc %>% display_top(5)
p515_cell_types <- update_cell_types(p515_cell_types, 5, cell_type = "monocyte")

# 6
rna_marker_list$p515_pbmc %>% display_top(6)
adt_marker_list$p515_pbmc %>% display_top(6)
one_v_six <- FindMarkers(pbmc_list$p515_pbmc, ident.1 = 1, ident.2 = 6, assay = "RNA")
p515_cell_types <- update_cell_types(p515_cell_types, 6, cell_type = "myeloid-like B cell")

# 7
rna_marker_list$p515_pbmc %>% display_top(7)
adt_marker_list$p515_pbmc %>% display_top(7)
# Compare to cluster 7
three_v_seven <- FindMarkers(pbmc_list$p515_pbmc, ident.1 = 3, ident.2 = 7, assay = "RNA")
p515_cell_types <- update_cell_types(p515_cell_types, 7, cell_type = "myeloid-like NK")

# 8
rna_marker_list$p515_pbmc %>% display_top(8)
adt_marker_list$p515_pbmc %>% display_top(8)
rna_marker_list$p515_pbmc %>% filter(cluster == 8 & gene == "CD1C")
p515_cell_types <- update_cell_types(p515_cell_types, 8, cell_type = "mDC")

# 9
rna_marker_list$p515_pbmc %>% display_top(9)
p515_cell_types <- update_cell_types(p515_cell_types, 9, cell_type = "T cell")

# 10
rna_marker_list$p515_pbmc %>% display_top(10)
adt_marker_list$p515_pbmc %>% display_top(10)
p515_cell_types <- update_cell_types(p515_cell_types, 10, cell_type = "T cell")

# Add cell types to Seurat object
p515_cell_type_vec <-
   setNames(as.character(p515_cell_types$cell_type),
          as.character(p515_cell_types$seurat_clusters))[as.character(pbmc_list$p515_pbmc$seurat_clusters)]
names(p515_cell_type_vec) <- NULL
pbmc_list$p515_pbmc$cell_type <- p515_cell_type_vec

# Add to COLORS vector
new_types <- p515_cell_type_vec %>% unique() %>% setdiff(names(COLORS))
n_add <- new_types %>% length()
COL_add <- my_colors[(length(COLORS) + 1): (length(COLORS) + n_add)]
names(COL_add) <- new_types
COLORS <- c(COLORS, COL_add)

# Plot it
s <- DimPlot(pbmc_list$p515_pbmc,
        reduction = "wnn.umap",
        group.by = "cell_type",
        pt.size = 2,
        alpha = 0.8,
        cols = COLORS, 
        combine = F)[[1]]
# Adjust labels
t <- s + labs(title = "Patient 515 annotated PBMC clusters") + xlab("UMAP coordinate 1") + ylab("UMAP coordinate 2")
# Take a look
t

# Save as PNG
umap_path <- paste0(project_path, "figures/")
# ggsave(paste0(umap_path, "p515_pbmc_umap", Sys.Date(), ".png"), t, height = 7, width = 9, bg = "white")

# Data frame to store cell types
p591_cell_types <-data.frame(seurat_clusters = levels(pbmc_list$p591_pbmc$seurat_clusters),
                         cell_type = NA)

# Take a look at 591 markers
# 0
rna_marker_list$p591_pbmc %>% display_top(0)
p591_cell_types <- update_cell_types(p591_cell_types, 0, cell_type = "monocyte")

# 1
rna_marker_list$p591_pbmc %>% display_top(1)
adt_marker_list$p591_pbmc %>% display_top(1)
FeaturePlot(pbmc_list$p591_pbmc, "CD14"); FeaturePlot(pbmc_list$p591_pbmc, "anti-human-CD14")
zero_v_one <- FindMarkers(pbmc_list$p591_pbmc, ident.1 = 0, ident.2 = 1, assay = "RNA")
# MNDA is a monocyte/M2 macrophage marker (in cluster 0, not 1)
# GOS2 is also a monocyte marker (in cluster 1, not 0)
# Given that they are all CD14+ and CD16-, they are classical monocytes
p591_cell_types <- update_cell_types(p591_cell_types, 1, cell_type = "monocyte")

# 2
rna_marker_list$p591_pbmc %>% display_top(2)
adt_marker_list$p591_pbmc %>% display_top(2)
p591_cell_types <- update_cell_types(p591_cell_types, 2, cell_type = "T cell")

# 3
rna_marker_list$p591_pbmc %>% display_top(3)
p591_cell_types <- update_cell_types(p591_cell_types, 3, cell_type = "B cell")

# 4
rna_marker_list$p591_pbmc %>% display_top(4)
p591_cell_types <- update_cell_types(p591_cell_types, 4, cell_type = "monocyte")

# 5
rna_marker_list$p591_pbmc %>% display_top(5)
adt_marker_list$p591_pbmc %>% display_top(5)
p591_cell_types <- update_cell_types(p591_cell_types, 5, cell_type = "natural killer")

# 6
rna_marker_list$p591_pbmc %>% display_top(6)
adt_marker_list$p591_pbmc %>% display_top(6)
p591_cell_types <- update_cell_types(p591_cell_types, 6, cell_type = "B cell")

# 7
rna_marker_list$p591_pbmc %>% display_top(7)
adt_marker_list$p591_pbmc %>% display_top(7)
p591_cell_types <- update_cell_types(p591_cell_types, 7, cell_type = "T cell")

# 8
rna_marker_list$p591_pbmc %>% display_top(8)
adt_marker_list$p591_pbmc %>% display_top(8)
p591_cell_types <- update_cell_types(p591_cell_types, 8, cell_type = "B cell")

# 9
rna_marker_list$p591_pbmc %>% display_top(9)
adt_marker_list$p591_pbmc %>% display_top(9)
p591_cell_types <- update_cell_types(p591_cell_types, 9, cell_type = "monocyte")

# 10
rna_marker_list$p591_pbmc %>% display_top(10)
adt_marker_list$p591_pbmc %>% display_top(10)
p591_cell_types <- update_cell_types(p591_cell_types, 10, cell_type = "mDC")

# 11
# CD34, MYB
rna_marker_list$p591_pbmc %>% display_top(11)
adt_marker_list$p591_pbmc %>% display_top(11)
rna_marker_list$p591_pbmc %>% filter(cluster == 11 & gene == "MYB")
p591_cell_types <- update_cell_types(p591_cell_types, 11, cell_type = "HSPC")

# 12
rna_marker_list$p591_pbmc %>% display_top(12)
p591_cell_types <- update_cell_types(p591_cell_types, 12, cell_type = "pDC")

# 13
rna_marker_list$p591_pbmc %>% display_top(13)
p591_cell_types <- update_cell_types(p591_cell_types, 13, cell_type = "platelets")

# Add cell types to Seurat object
p591_cell_type_vec <-
   setNames(as.character(p591_cell_types$cell_type),
          as.character(p591_cell_types$seurat_clusters))[as.character(pbmc_list$p591_pbmc$seurat_clusters)]
names(p591_cell_type_vec) <- NULL
pbmc_list$p591_pbmc$cell_type <- p591_cell_type_vec

# Add to COLORS vector
new_types <- p591_cell_type_vec %>% unique() %>% setdiff(names(COLORS))
n_add <- new_types %>% length()
COL_add <- my_colors[(length(COLORS) + 1): (length(COLORS) + n_add)]
names(COL_add) <- new_types
COLORS <- c(COLORS, COL_add)

# Plot it
s <- DimPlot(pbmc_list$p591_pbmc,
        reduction = "wnn.umap",
        group.by = "cell_type",
        pt.size = 2,
        alpha = 0.8,
        cols = COLORS, 
        combine = F)[[1]]
# Adjust labels
t <- s + labs(title = "Patient 591 annotated PBMC clusters") + xlab("UMAP coordinate 1") + ylab("UMAP coordinate 2")
# Take a look
t

# Save as PNG
umap_path <- paste0(project_path, "figures/")
ggsave(paste0(umap_path, "p591_pbmc_umap", Sys.Date(), ".png"), t, height = 7, width = 9, bg = "white")

# Data frame to store cell types
p594_cell_types <-data.frame(seurat_clusters = levels(pbmc_list$p594_pbmc$seurat_clusters),
                         cell_type = NA)

# Take a look at 594 markers

# 0
rna_marker_list$p594_pbmc %>% display_top(0)
p594_cell_types <- update_cell_types(p594_cell_types, 0, cell_type = "monocyte")

# 1
rna_marker_list$p594_pbmc %>% display_top(1)
p594_cell_types <- update_cell_types(p594_cell_types, 1, cell_type = "B cell")

# 2
rna_marker_list$p594_pbmc %>% display_top(2)
p594_cell_types <- update_cell_types(p594_cell_types, 2, cell_type = "T cell")

# 3
rna_marker_list$p594_pbmc %>% display_top(3)
adt_marker_list$p594_pbmc %>% display_top(3)
p594_cell_types <- update_cell_types(p594_cell_types, 3, cell_type = "natural killer")

# 4
rna_marker_list$p594_pbmc %>% display_top(4)
p594_cell_types <- update_cell_types(p594_cell_types, 4, cell_type = "monocyte")

# 5
rna_marker_list$p594_pbmc %>% display_top(5)
adt_marker_list$p594_pbmc %>% display_top(5)
p594_cell_types <- update_cell_types(p594_cell_types, 5, cell_type = "monocyte")

# 6
rna_marker_list$p594_pbmc %>% display_top(6)
p594_cell_types <- update_cell_types(p594_cell_types, 6, cell_type = "T cell")

# 7
rna_marker_list$p594_pbmc %>% display_top(7)
p594_cell_types <- update_cell_types(p594_cell_types, 7, cell_type = "mDC")

# 8 (biphenotypic, B and C)
rna_marker_list$p594_pbmc %>% display_top(8)
adt_marker_list$p594_pbmc %>% display_top(8)
p594_cell_types <- update_cell_types(p594_cell_types, 8, cell_type = "monocyte")

# 9
rna_marker_list$p594_pbmc %>% display_top(9)
p594_cell_types <- update_cell_types(p594_cell_types, 9, cell_type = "pDC")


# Add cell types to Seurat object
p594_cell_type_vec <-
   setNames(as.character(p594_cell_types$cell_type),
          as.character(p594_cell_types$seurat_clusters))[as.character(pbmc_list$p594_pbmc$seurat_clusters)]
names(p594_cell_type_vec) <- NULL
pbmc_list$p594_pbmc$cell_type <- p594_cell_type_vec

# Add to COLORS vector
new_types <- p594_cell_type_vec %>% unique() %>% setdiff(names(COLORS))
n_add <- new_types[!is.na(new_types)] %>% length()
if (n_add > 0) {
  COL_add <- my_colors[(length(COLORS) + 1): (length(COLORS) + n_add)]
  names(COL_add) <- new_types
  COLORS <- c(COLORS, COL_add)
}

# Plot it
s <- DimPlot(pbmc_list$p594_pbmc,
        reduction = "wnn.umap",
        group.by = "cell_type",
        pt.size = 2,
        alpha = 0.8,
        cols = COLORS, 
        combine = F)[[1]]
# Adjust labels
t <- s + labs(title = "Patient 594 annotated PBMC clusters") + xlab("UMAP coordinate 1") + ylab("UMAP coordinate 2")
# Take a look
t

# Save as PNG
umap_path <- paste0(project_path, "figures/")
ggsave(paste0(umap_path, "p594_pbmc_umap", Sys.Date(), ".png"), t, height = 7, width = 9, bg = "white")
```

```{r}
plot_pb_offset(pbo_monocytes, )
```

