---
title: "jSSc ASCT monocytes longitudinal analysis"
output: html_notebook
---

In this notebook, we run our longitudinal monocyte pseudobulk offset analysis. We use a single model which includes an indicator term for tissue. 

We begin by loading in useful packages and custom functions.

First, clear the environment and set the path where the project folder is saved.

```{r}
# Clear all objects from environment
rm(list = ls(all.names = TRUE))
# Free up memory
gc()
# Increase max vector memory
mem.maxVSize(200000)

project_path <- "/Users/juliaelrod/Desktop/cmu/research/scleroderma/jSSc_ASCT_monocytes_manuscript/code_for_manuscript/"
```

Now load in useful packages and custom functions.

```{r}
# Load packages
packages <- c("Seurat", "dplyr", "tidyr", "SeuratDisk", "harmony", "readr",
              "ggplot2", "stringr", "SoupX", "dsb", "data.table", "Matrix",
              "scDblFinder", "scater", "purrr")
invisible(lapply(packages, require, character.only = TRUE))

# Source functions
function_files <- list.files(paste0(project_path, "functions"), pattern = "\\.R$", full.names = TRUE)
invisible(lapply(function_files, source))
```

Now, we load in the most recent Seurat objects for PBMCs and skin, subset out monocytes and macrophages (the focus of our manuscript), and store the data in a single Seurat object. We can only include genes expressed in both tissues, leaving us with 17,836 genes.

```{r}
# # Load in PBMC Seurat object
# scler_pbmc <- LoadH5Seurat(paste0(project_path, "data/h5seurat/scler_pbmc_10.h5seurat"))
# 
# # Subset out  monocytes, separately for SSc and healthy.
# healthy_monocytes <- subset(scler_pbmc, cell_type %in% c("CD14+ monocyte", "CD16+ monocyte") & health == "healthy")
# SSc_monocytes <- subset(scler_pbmc, cell_type %in% c("CD14+ monocyte", "CD16+ monocyte") & health == "SSc")
# 
# # Sub-cluster separately for SSc and healthy
# healthy_monocytes <- healthy_monocytes %>% sub_cluster(res = 0.1, dim = c(5, 5))
# SSc_monocytes <- SSc_monocytes %>% sub_cluster(res = 1)
# 
# # Save as H5
# SaveH5Seurat(healthy_monocytes, paste0(project_path, "data/h5seurat/healthy_monocytes"))
# SaveH5Seurat(SSc_monocytes, paste0(project_path, "data/h5seurat/SSc_monocytes"))
# 
# # Load in healthy monocytes
# healthy_monocytes <- LoadH5Seurat(paste0(project_path, "data/h5seurat/healthy_monocytes.h5seurat"))
# SSc_monocytes <- LoadH5Seurat(paste0(project_path, "data/h5seurat/SSc_monocytes.h5seurat"))
# 
# # List of assays
# ASSAYS <- c("RNA", "ADT")
# 
# # Get markers for healthy controls
# healthy_markers <- ASSAYS %>% lapply(function(x){FindAllMarkers(healthy_monocytes, assay = x, only.pos = T)}) %>% setNames(ASSAYS)
# 
# # Annotate healthy monocytes
# healthy_markers %>% lapply(function(x){lapply(unique(x$cluster),
#                                               function(clus){x %>% filter(cluster == clus) %>% head(20) %>% pull(gene)}) %>% setNames(paste0("Cluster_", unique(x$cluster)))})
# 
# # Cluster 0 is CD14+, cluster 1 is CD16+
# healthy_monocytes@meta.data <- healthy_monocytes@meta.data %>%
#   mutate(
#     cell_type = case_when(
#       seurat_clusters == 0 ~ "CD14+",
#       seurat_clusters == 1 ~ "CD16+"
#     )
#   )
# 
# # Save healthy annotations
# SaveH5Seurat(healthy_monocytes, paste0(project_path, "data/h5seurat/healthy_monocytes_2"))

# # Load in annotated healthy monoytes
# healthy_monocytes <- LoadH5Seurat(paste0(project_path, "data/h5seurat/healthy_monocytes_2.h5seurat"))
# 
# # Take a look
# DimPlot(healthy_monocytes, reduction = "wnn.umap", label = T, group.by = "cell_type", label.size = 5, label.box = T) + ggtitle("Healthy monocytes")

# # Load in SSc monocytes
# SSc_monocytes <- LoadH5Seurat(paste0(project_path, "data/h5seurat/SSc_monocytes.h5seurat"))
#
# # Now, let's sub-cluster separately for each patient
#
# # Subset for each patient
# monocytes_515 <- subset(SSc_monocytes, subject == "515")
# monocytes_591 <- subset(SSc_monocytes, subject == "591")
# monocytes_594 <- subset(SSc_monocytes, subject == "594")
# 
# # Sub-cluster each patient
# monocytes_515 <- monocytes_515 %>% sub_cluster(res = 1)
# monocytes_591 <- monocytes_591 %>% sub_cluster(res = 1)
# monocytes_594 <- monocytes_594 %>% sub_cluster(res = 1)
# 
# # Save as H5
# SaveH5Seurat(monocytes_515, paste0(project_path, "data/h5seurat/monocytes_515"))
# SaveH5Seurat(monocytes_591, paste0(project_path, "data/h5seurat/monocytes_591"))
# SaveH5Seurat(monocytes_594, paste0(project_path, "data/h5seurat/monocytes_594"))
# 
# # Load in H5s for each individual patient
# monocytes_515 <- LoadH5Seurat(paste0(project_path, "data/h5seurat/monocytes_515.h5seurat"))
# monocytes_591 <- LoadH5Seurat(paste0(project_path, "data/h5seurat/monocytes_591.h5seurat"))
# monocytes_594 <- LoadH5Seurat(paste0(project_path, "data/h5seurat/monocytes_594.h5seurat"))
# 
# # patient 515--CD14, CD16, and B-like
# DimPlot(monocytes_515, reduction = "wnn.umap", label = T) + ggtitle("Patient 515 monocytes")
# DotPlot(monocytes_515, c("CD14", "FCGR3A", "MS4A1"))
# monocytes_515@meta.data <- monocytes_515@meta.data %>% mutate(
#   cell_type = case_when(
#     seurat_clusters %in% c(0, 1, 2, 4, 5, 6, 7) ~ "CD14+",
#     seurat_clusters == 3 ~ "CD16+",
#     seurat_clusters == 8 ~ "B-like"
#   )
# )
# 
# # patient 591 - CD14, CD16, B-like, T-like, NK-like
# DimPlot(monocytes_591, reduction = "wnn.umap", label = T) + ggtitle("Patient 591 monocytes")
# DotPlot(monocytes_591, c("CD14", "FCGR3A", "MS4A1", "CD3D", "NKG7")) + ggtitle("Patient 591 monocytes")
# monocytes_591@meta.data <- monocytes_591@meta.data %>% mutate(
#   cell_type = case_when(
#     seurat_clusters %in% c(0, 1, 2, 3, 4, 6, 8, 12) ~ "CD14+",
#     seurat_clusters == 5 ~ "Intermediate",
#     seurat_clusters %in% c(7, 11) ~ "CD16+",
#     seurat_clusters == 9 ~ "B-like",
#     seurat_clusters == 10 ~ "T-like",
#     seurat_clusters == 13 ~ "NK-like"
#   )
# )
# 
# # patient 594 - CD14, CD16
# DimPlot(monocytes_594, reduction = "wnn.umap", label = T) + ggtitle("Patient 594 monocytes")
# DotPlot(monocytes_594, c("CD14", "FCGR3A")) + ggtitle("Patient 594 monocytes")
# monocytes_594@meta.data <- monocytes_594@meta.data %>% mutate(
#   cell_type = case_when(
#     seurat_clusters %in% c(0, 1, 2, 4, 5, 6) ~ "CD14+",
#     seurat_clusters == 3 ~ "CD16+",
#   )
# )
# 
# # Save annotated patient monocytes as H5
# SaveH5Seurat(monocytes_515, paste0(project_path, "data/h5seurat/monocytes_515_2.h5seurat"))
# SaveH5Seurat(monocytes_591, paste0(project_path, "data/h5seurat/monocytes_591_2.h5seurat"))
# SaveH5Seurat(monocytes_594, paste0(project_path, "data/h5seurat/monocytes_594_2.h5seurat"))


# # Load in annotated monocytes by patient
# monocytes_515 <- LoadH5Seurat(paste0(project_path, "data/h5seurat/monocytes_515_2.h5seurat"))
# monocytes_591 <- LoadH5Seurat(paste0(project_path, "data/h5seurat/monocytes_591_2.h5seurat"))
# monocytes_594 <- LoadH5Seurat(paste0(project_path, "data/h5seurat/monocytes_594_2.h5seurat"))
# 
# # Function to plot cell type proportions
# plot_cell_type_prop <- function(CELLTYPE, seurat_obj) {
#   ct_dat <- seurat_obj@meta.data %>% filter(cell_type == CELLTYPE) %>% pull(timepoint) %>% table() %>% as.data.frame()
#   colnames(ct_dat) <- c("time", "count")
#   tot_dat <- seurat_obj@meta.data %>% pull(timepoint) %>% table() %>% as.data.frame()
#   colnames(tot_dat) <- c("time", "count")
#   prop_dat <- data.frame(
#     time = ct_dat$time,
#     prop = (ct_dat$count / tot_dat[tot_dat$time %in% ct_dat$time, ]$count)
#     )
#   ggplot(prop_dat, aes(y = prop, x = as.numeric(as.character(time)) , group = 1)) +
#     geom_line() + 
#     theme_bw() +
#     ggtitle(paste0("Patient 515: ", CELLTYPE, " monocytes")) +
#     xlab("Months since transplant") +
#     ylab("Proportion of all monocytes")
#   }
# 
# # We plot cell type proportions
# 
# monocytes_515_cell_prop_plots <- unique(monocytes_515$cell_type) %>%
#   lapply(plot_cell_type_prop, monocytes_515) %>%
#   setNames(unique(monocytes_515$cell_type))
# 
# monocytes_591_cell_prop_plots <- unique(monocytes_591$cell_type) %>%
#   lapply(plot_cell_type_prop, monocytes_591) %>%
#   setNames(unique(monocytes_591$cell_type))
# 
# monocytes_594_cell_prop_plots <- unique(monocytes_594$cell_type) %>%
#   lapply(plot_cell_type_prop, monocytes_594) %>%
#   setNames(unique(monocytes_594$cell_type))
# 
# # We also make dim plots
# DimPlot(monocytes_515, reduction = "wnn.umap", group.by = "cell_type", label = T, repel = F) + ggtitle("Patient 515 monocytes")
# DimPlot(monocytes_591, reduction = "wnn.umap", group.by = "cell_type", label = T, repel = F) + ggtitle("Patient 591 monocytes")
# DimPlot(monocytes_594, reduction = "wnn.umap", group.by = "cell_type", label = T, repel = F) + ggtitle("Patient 594 monocytes")

# # Now, we add these cell types back to the SSc_monocytes object
# # Build one combined lookup vector
# 
# celltype_lookup <- c(
#   setNames(monocytes_515$cell_type, colnames(monocytes_515)),
#   setNames(monocytes_591$cell_type, colnames(monocytes_591)),
#   setNames(monocytes_594$cell_type, colnames(monocytes_594))
# )
# 
# # Map directly by cell_ID
# SSc_monocytes$cell_type <- celltype_lookup[SSc_monocytes$cell_ID]
# 
# # Save annotated SSc_monocytes
# SaveH5Seurat(SSc_monocytes, paste0(project_path, "data/h5seurat/SSc_monocytes_2.h5seurat"))
```

Now, we run the model on all the patient monocytes, with an indicator for cell type. 

```{r}
# Read in data
SSc_monocytes <- LoadH5Seurat(paste0(project_path, "data/h5seurat/SSc_monocytes_2.h5seurat"))

# First, try with indicator for cell type

# Design formula for our NB GLM. Covariates are subject (indicator), time, and cell_type (indicator)
# Center and scale and numeric variables (VERY IMPORTANT--not doing so can lead to numeric instability, missing significant results, etc.)
# poly() effectively does this for us.
design_formula <- ~ subject + poly(timepoint) * cell_type

# Create a new variable that is cell_type and sample_id combined
SSc_monocytes$sample_id_cell_type <- paste0(SSc_monocytes$sample_id,"_", SSc_monocytes$cell_type)

# Fit offset pseudobulk model
pbo_monocytes <- pb_offset(
  seurat_obj = SSc_monocytes,
  sample_id_variable = "sample_id_cell_type",
  design_formula = design_formula)

# Test time coefficient for B-like cells
B_like_results <- test_de(pbo_monocytes$fit, `poly(timepoint)`) %>% arrange(pval) %>% filter(pval < 0.1)
# Test time coefficient for CD14+ cells
CD14_results <- test_de(pbo_monocytes$fit, `poly(timepoint)` + `poly(timepoint):cell_typeCD14+`) %>% arrange(pval) %>% filter(pval < 0.1)
# Test time coefficient for CD16+ cells
CD16_results <- test_de(pbo_monocytes$fit, `poly(timepoint)` + `poly(timepoint):cell_typeCD16+`) %>% arrange(pval) %>% filter(pval < 0.1)
# Test time coefficient for Intermediate cells
Intermediate_results <- test_de(pbo_monocytes$fit, `poly(timepoint)` + `poly(timepoint):cell_typeIntermediate`) %>% arrange(pval) %>% filter(pval < 0.1)
# Test time coefficient for NK-like cells
NK_like_results <- test_de(pbo_monocytes$fit, `poly(timepoint)` + `poly(timepoint):cell_typeNK-like`) %>% arrange(pval) %>% filter(pval < 0.1)
# Test time coefficient for T-like cells
T_like_results <- test_de(pbo_monocytes$fit, `poly(timepoint)` + `poly(timepoint):cell_typeT-like`) %>% arrange(pval) %>% filter(pval < 0.1)

# Pop into a list
results_list <- list(B_like = B_like_results,
                  CD14 = CD14_results,
                  CD16 = CD16_results,
                  Intermediate = Intermediate_results,
                  NK_like = NK_like_results,
                  T_like = T_like_results)

# Extract significant DEGs
DEGs <- results_list %>% map("name")

# How many DEGs per cell type?
DEGs %>% lapply(length) 
```

Now we conduct pathway enrichment analysis based off of this tutorial:  https://biostatsquid.com/pathway-enrichment-analysis-tutorial-clusterprofiler/

Setup for GSEA is below:

```{r}
# Load packages
packages_2 <- c("tidyverse", "RColorBrewer", "pheatmap", "clusterProfiler", "enrichplot", "msigdbr")
invisible(lapply(packages_2, require, character.only = TRUE))

# Get hallmark, GO biological process, and GO cellular component pathways
hallmark_sets <- msigdbr(species = "human", 
                         collection = "H")
bp_sets <- msigdbr(species = "human",
                   collection = "C5",
                   subcollection = "GO:BP")
cc_sets <- msigdbr(species = "human",
                   collection = "C5",
                   subcollection = "GO:CC")
# Store all sets together
all_sets <- rbind(hallmark_sets, bp_sets, cc_sets)

# TERM2GENE
term2gene <- all_sets %>%
  select(gs_name, gene_symbol)

# TERM2NAME (readable descriptions)
term2name <- all_sets %>%
  distinct(gs_name, gs_description)

# Background genes
DefaultAssay(SSc_monocytes) <- "RNA"
background_genes <- rownames(SSc_monocytes)
```

Let's run pathway analysis for each tissue separately.

```{r}
# (1) PBMC

# Add direction (whether expression goes up or down over time) to results data frame
results_list <- results_list %>% lapply(function(x){
  x %>%
  mutate(direction = case_when(
  lfc > 0 ~ 'UP',
  lfc < 0 ~ 'DOWN'
  ))
})

# Split each dataframe into a list of sub-dataframes: upregulated, downregulated genes
results_nested_list <- results_list %>% lapply(function(x){
  split(x, x$direction)
})

# Run enricher R on each set of genes (for each cell tyep: up set and down set)
gsea <- map(results_nested_list, ~
  map(.x, ~
    enricher(
      gene = .x$name,
      TERM2GENE = term2gene,
      TERM2NAME = term2name,
      universe = background_genes,
      pvalueCutoff = 1,
      qvalueCutoff = 1,
      minGSSize = 5,
      maxGSSize = 500
    )
  )
)

```

Now, we take a look at the results

```{r}
# Downward trends by cell type
gsea %>% lapply(function(x){x$DOWN@result %>% head() %>% select(p.adjust, Count)}) 

# Upward trends by cell type
gsea %>% lapply(function(x){x$UP@result %>% head() %>% select(p.adjust, Count)}) 

```

```{r}
# Path to save results
results_dir <- paste0(project_path, "../to_share/results_03_04_26/by_subtype/")

names(results_nested_list) %>% lapply(function(x){DIR <- paste0(results_dir, x)
if (!dir.exists(DIR)) {dir.create(DIR)}
results_nested_list[[x]]$DOWN %>% write_csv(paste0(DIR, "/", x, "_down_DEGs.csv"))
results_nested_list[[x]]$UP %>% write_csv(paste0(DIR, "/", x, "_up_DEGs.csv"))
gsea[[x]]$DOWN@result %>% write_csv(paste0(DIR, "/", x, "_down_GSEA.csv"))
gsea[[x]]$UP@result %>% write_csv(paste0(DIR, "/", x, "_up_GSEA.csv"))
})
```
