---
title: "4: jSSc ASCT monocytes longitudinal analysis"
output: html_notebook
---

In this notebook, we run our longitudinal monocyte pseudobulk offset analysis. We use a single model which includes an indicator term for tissue. 

We begin by loading in useful packages and custom functions.

First, clear the environment and set the path where the project folder is saved.

```{r}
# Clear all objects from environment
rm(list = ls(all.names = TRUE))
# Free up memory
gc()
# Increase max vector memory
mem.maxVSize(200000)

project_path <- "/Users/juliaelrod/Desktop/cmu/research/scleroderma/jSSc_ASCT_monocytes_manuscript/code_for_manuscript/"

results_dir <- paste0(project_path, "data/DEG_pathway_CSVs/")
```

Now load in useful packages and custom functions.

```{r}
# Load packages
packages <- c("Seurat", "dplyr", "tidyr", "SeuratDisk", "harmony", "readr",
              "ggplot2", "stringr", "SoupX", "dsb", "data.table", "Matrix",
              "scDblFinder", "scater", "ggrepel")
invisible(lapply(packages, require, character.only = TRUE))

# Source functions
function_files <- list.files(paste0(project_path, "functions"), pattern = "\\.R$", full.names = TRUE)
invisible(lapply(function_files, source))
```

Now, we load in the most recent Seurat objects for PBMCs and skin, subset out monocytes, and store the data in a single Seurat object. 

```{r}
# # Load in PBMC Seurat object
# scler_pbmc <- LoadH5Seurat(paste0(project_path, "data/h5seurat/scler_pbmc_10.h5seurat"))
# 
# # Subset out  monocytes, separately for SSc and healthy.
# healthy_monocytes <- subset(scler_pbmc, cell_type == "monocyte" & health == "healthy")
# SSc_monocytes <- subset(scler_pbmc, cell_type == "monocyte" & health == "SSc")
# 
# # Sub-cluster separately for SSc and healthy
# healthy_monocytes <- healthy_monocytes %>% sub_cluster(res = 0.1, dim = c(5, 5))
# SSc_monocytes <- SSc_monocytes %>% sub_cluster(res = 1)
# 
# # Save as H5
# SaveH5Seurat(healthy_monocytes, paste0(project_path, "data/h5seurat/healthy_monocytes"))
# SaveH5Seurat(SSc_monocytes, paste0(project_path, "data/h5seurat/SSc_monocytes"))
# 
# # Load in healthy monocytes
# healthy_monocytes <- LoadH5Seurat(paste0(project_path, "data/h5seurat/healthy_monocytes.h5seurat"))
# SSc_monocytes <- LoadH5Seurat(paste0(project_path, "data/h5seurat/SSc_monocytes.h5seurat"))
#
# # List of assays
# ASSAYS <- c("RNA", "ADT")
# 
# # Get markers for healthy controls
# healthy_markers <- ASSAYS %>% lapply(function(x){FindAllMarkers(healthy_monocytes, assay = x, only.pos = T)}) %>% setNames(ASSAYS)
# 
# # Annotate healthy monocytes
# healthy_markers %>% lapply(function(x){lapply(unique(x$cluster),
#                                               function(clus){x %>% filter(cluster == clus) %>% head(20) %>% pull(gene)}) %>% setNames(paste0("Cluster_", unique(x$cluster)))})
# 
# # Cluster 0 is CD14+, cluster 1 is CD16+
# healthy_monocytes@meta.data <- healthy_monocytes@meta.data %>%
#   mutate(
#     cell_type = case_when(
#       seurat_clusters == 0 ~ "CD14+",
#       seurat_clusters == 1 ~ "CD16+"
#     )
#   )
# 
# # Save healthy annotations
# SaveH5Seurat(healthy_monocytes, paste0(project_path, "data/h5seurat/healthy_monocytes_2"))
# 
# # Load in annotated healthy monoytes
# healthy_monocytes <- LoadH5Seurat(paste0(project_path, "data/h5seurat/healthy_monocytes_2.h5seurat"))
# 
# # Take a look
# DimPlot(healthy_monocytes, reduction = "wnn.umap", label = T, group.by = "cell_type", label.size = 5, label.box = T) + ggtitle("Healthy monocytes")
# 
# Load in SSc monocytes
SSc_monocytes <- LoadH5Seurat(paste0(project_path, "data/h5seurat/SSc_monocytes.h5seurat"))
#
# # Now, let's sub-cluster separately for each patient
#
# # Subset for each patient
# monocytes_515 <- subset(SSc_monocytes, subject == "515")
# monocytes_591 <- subset(SSc_monocytes, subject == "591")
# monocytes_594 <- subset(SSc_monocytes, subject == "594")
# 
# # Sub-cluster each patient
# monocytes_515 <- monocytes_515 %>% sub_cluster(res = 1)
# monocytes_591 <- monocytes_591 %>% sub_cluster(res = 1)
# monocytes_594 <- monocytes_594 %>% sub_cluster(res = 1)
# 
# # Save as H5
# SaveH5Seurat(monocytes_515, paste0(project_path, "data/h5seurat/monocytes_515"))
# SaveH5Seurat(monocytes_591, paste0(project_path, "data/h5seurat/monocytes_591"))
# SaveH5Seurat(monocytes_594, paste0(project_path, "data/h5seurat/monocytes_594"))
# 
# # Load in H5s for each individual patient
# monocytes_515 <- LoadH5Seurat(paste0(project_path, "data/h5seurat/monocytes_515.h5seurat"))
# monocytes_591 <- LoadH5Seurat(paste0(project_path, "data/h5seurat/monocytes_591.h5seurat"))
# monocytes_594 <- LoadH5Seurat(paste0(project_path, "data/h5seurat/monocytes_594.h5seurat"))
# 
# # patient 515--CD14, CD16, and B-like
# DimPlot(monocytes_515, reduction = "wnn.umap", label = T) + ggtitle("Patient 515 monocytes")
# DotPlot(monocytes_515, c("CD14", "FCGR3A", "MS4A1"))
# monocytes_515@meta.data <- monocytes_515@meta.data %>% mutate(
#   cell_type = case_when(
#     seurat_clusters %in% c(0, 1, 2, 4, 5, 6, 7) ~ "CD14+",
#     seurat_clusters == 3 ~ "CD16+",
#     seurat_clusters == 8 ~ "B-like"
#   )
# )
# 
# # patient 591 - CD14, CD16, B-like, T-like, NK-like
# DimPlot(monocytes_591, reduction = "wnn.umap", label = T) + ggtitle("Patient 591 monocytes")
# DotPlot(monocytes_591, c("CD14", "FCGR3A", "MS4A1", "CD3D", "NKG7")) + ggtitle("Patient 591 monocytes")
# monocytes_591@meta.data <- monocytes_591@meta.data %>% mutate(
#   cell_type = case_when(
#     seurat_clusters %in% c(0, 1, 2, 3, 4, 6, 8, 12) ~ "CD14+",
#     seurat_clusters == 5 ~ "Intermediate",
#     seurat_clusters %in% c(7, 11) ~ "CD16+",
#     seurat_clusters == 9 ~ "B-like",
#     seurat_clusters == 10 ~ "T-like",
#     seurat_clusters == 13 ~ "NK-like"
#   )
# )
# 
# # patient 594 - CD14, CD16
# DimPlot(monocytes_594, reduction = "wnn.umap", label = T) + ggtitle("Patient 594 monocytes")
# DotPlot(monocytes_594, c("CD14", "FCGR3A")) + ggtitle("Patient 594 monocytes")
# monocytes_594@meta.data <- monocytes_594@meta.data %>% mutate(
#   cell_type = case_when(
#     seurat_clusters %in% c(0, 1, 2, 4, 5, 6) ~ "CD14+",
#     seurat_clusters == 3 ~ "CD16+",
#   )
# )
# 
# # Save annotated patient monocytes as H5
# SaveH5Seurat(monocytes_515, paste0(project_path, "data/h5seurat/monocytes_515_2.h5seurat"))
# SaveH5Seurat(monocytes_591, paste0(project_path, "data/h5seurat/monocytes_591_2.h5seurat"))
# SaveH5Seurat(monocytes_594, paste0(project_path, "data/h5seurat/monocytes_594_2.h5seurat"))
# 
# 
# # Load in annotated monocytes by patient
# monocytes_515 <- LoadH5Seurat(paste0(project_path, "data/h5seurat/monocytes_515_2.h5seurat"))
# monocytes_591 <- LoadH5Seurat(paste0(project_path, "data/h5seurat/monocytes_591_2.h5seurat"))
# monocytes_594 <- LoadH5Seurat(paste0(project_path, "data/h5seurat/monocytes_594_2.h5seurat"))
# 
# # Function to plot cell type proportions
# plot_cell_type_prop <- function(CELLTYPE, seurat_obj) {
#   ct_dat <- seurat_obj@meta.data %>% filter(cell_type == CELLTYPE) %>% pull(timepoint) %>% table() %>% as.data.frame()
#   colnames(ct_dat) <- c("time", "count")
#   tot_dat <- seurat_obj@meta.data %>% pull(timepoint) %>% table() %>% as.data.frame()
#   colnames(tot_dat) <- c("time", "count")
#   prop_dat <- data.frame(
#     time = ct_dat$time,
#     prop = (ct_dat$count / tot_dat[tot_dat$time %in% ct_dat$time, ]$count)
#     )
#   ggplot(prop_dat, aes(y = prop, x = as.numeric(as.character(time)) , group = 1)) +
#     geom_line() + 
#     theme_bw() +
#     ggtitle(paste0("Patient 515: ", CELLTYPE, " monocytes")) +
#     xlab("Months since transplant") +
#     ylab("Proportion of all monocytes")
#   }
# 
# # We plot cell type proportions
# 
# monocytes_515_cell_prop_plots <- unique(monocytes_515$cell_type) %>%
#   lapply(plot_cell_type_prop, monocytes_515) %>%
#   setNames(unique(monocytes_515$cell_type))
# 
# monocytes_591_cell_prop_plots <- unique(monocytes_591$cell_type) %>%
#   lapply(plot_cell_type_prop, monocytes_591) %>%
#   setNames(unique(monocytes_591$cell_type))
# 
# monocytes_594_cell_prop_plots <- unique(monocytes_594$cell_type) %>%
#   lapply(plot_cell_type_prop, monocytes_594) %>%
#   setNames(unique(monocytes_594$cell_type))
# 
# # We also make dim plots
# DimPlot(monocytes_515, reduction = "wnn.umap", group.by = "cell_type", label = T, repel = F) + ggtitle("Patient 515 monocytes")
# DimPlot(monocytes_591, reduction = "wnn.umap", group.by = "cell_type", label = T, repel = F) + ggtitle("Patient 591 monocytes")
# DimPlot(monocytes_594, reduction = "wnn.umap", group.by = "cell_type", label = T, repel = F) + ggtitle("Patient 594 monocytes")

# # Now, we add these cell types back to the SSc_monocytes object
# # Build one combined lookup vector
# 
# celltype_lookup <- c(
#   setNames(monocytes_515$cell_type, colnames(monocytes_515)),
#   setNames(monocytes_591$cell_type, colnames(monocytes_591)),
#   setNames(monocytes_594$cell_type, colnames(monocytes_594))
# )
# 
# # Map directly by cell_ID
# SSc_monocytes$cell_type <- celltype_lookup[SSc_monocytes$cell_ID]
# 
# # Save annotated SSc_monocytes
# SaveH5Seurat(SSc_monocytes, paste0(project_path, "data/h5seurat/SSc_monocytes_2.h5seurat"))
```

Now, we run the model on all the patient monocytes. 

```{r}
# Read in data
SSc_monocytes <- LoadH5Seurat(paste0(project_path, "data/h5seurat/SSc_monocytes_2.h5seurat"))

# Design
design_formula <- ~ subject + poly(timepoint)

# Fit offset pseudobulk model
pbo_monocytes <- pb_offset(
  seurat_obj = SSc_monocytes,
  sample_id_variable = "sample_id",
  design_formula = design_formula)

# Test for significant effect of time since ASCT
results <- test_de(pbo_monocytes$fit, `poly(timepoint)`) %>% 
  arrange(pval)

# Save all results
write_csv(results, paste0(results_dir, "all_DE_results.csv"))
results <- read_csv(paste0(results_dir, "all_DE_results.csv"))
```

Volcano plot of top DEGs:

First, we grab genes from fav up pathways

```{r}
# Get all gene sets
gene_sets <- gsea$DOWN@geneSets
# Extract 5 favorite pathways
fav_gene_sets <- gene_sets[favorite_pathways]

# Make a named vector of p-values for quick lookup
pval_lookup <- results$pval
names(pval_lookup) <- results$name

# Order each pathway's genes by p-value
fav_gene_sets_ordered <- lapply(fav_gene_sets, function(genes) {
  # Keep only genes that exist in results
  genes_in_results <- intersect(genes, names(pval_lookup))
  
  # Order by p-value
  genes_in_results[order(pval_lookup[genes_in_results])]
})

topgenes <- lapply(fav_gene_sets_ordered, function(x){head(x, 5)}) %>% unlist() %>% setNames(NULL) %>% unique()
```


```{r}
results <- read_csv(paste0(results_dir, "all_DE_results.csv"))

top10_up <- results %>% arrange(pval) %>% filter(lfc > 0.6) %>% pull(name) %>% head(10)

top10_down <- results %>% arrange(pval) %>% filter(lfc < -0.6) %>% pull(name) %>% head(10)

volc_down_degs <- levels(heat_df$gene)

volc_up_degs <- results %>% arrange(pval) %>% filter(lfc < -0.6) %>% pull(name) %>% head(18)

results <- results %>%
  dplyr::rename(log2fc = lfc) %>%
  mutate(`Trend following ASCT` = case_when(
    log2fc > 0.6 & pval < 0.05 ~ "Up",
    log2fc < -0.6 & pval < 0.05 ~ "Down",
    TRUE ~ "Not significant"
  )) %>%
  mutate(delabel = case_when(
    name %in% c(volc_down_degs, volc_up_degs) ~ name,
    TRUE ~ NA
  ))

results$`Trend following ASCT` <- factor(results$`Trend following ASCT`, levels = c("Up", "Down", "Not significant"))

volcanoplot <- ggplot(data = results, aes(x = log2fc, y = -log10(pval), col = `Trend following ASCT`, label = delabel)) +
  geom_vline(xintercept = c(-0.6, 0.6), col = "gray", linetype = 'dashed') +
  geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed') + 
  geom_point(size = 2) +
  theme_classic(base_size = 20) +
  theme(axis.title.y = element_text(face = "bold", margin = margin(0,20,0,0), size = rel(1.1), color = 'black'),
        axis.title.x = element_text(hjust = 0.5, face = "bold", margin = margin(20,0,0,0), size = rel(1.1), color = 'black'),
        plot.title = element_text(hjust = 0.5, size = rel(1.1))) +
  scale_color_manual(values = c("red3", "#00AFBB", "grey")) +
  labs(x = expression("log"[2]*"FC"), y = expression("-log"[10]*"p-value")) + 
  scale_x_continuous(breaks = seq(-10, 10, 2)) +
  ggtitle("Differentially expressed genes over time \nsince ASCT in jSSc monocytes") +
  geom_label_repel(max.overlaps = Inf,
                  size = rel(3),
                  box.padding = 0.3,
                  force = 4,
                  show.legend = F)

# Save as PNG
ggsave(paste0(project_path, "figures/volcano_plot.png"), volcanoplot, width = 12, height = 6)
```

Now we conduct pathway enrichment analysis based off of this tutorial:  https://biostatsquid.com/pathway-enrichment-analysis-tutorial-clusterprofiler/

Setup for GSEA is below:

```{r}
# Load packages
packages_2 <- c("tidyverse", "RColorBrewer", "pheatmap", "clusterProfiler", "enrichplot", "msigdbr")
invisible(lapply(packages_2, require, character.only = TRUE))

# Get hallmark, GO biological process, and GO cellular component pathways
hallmark_sets <- msigdbr(species = "human",
                         collection = "H")
bp_sets <- msigdbr(species = "human",
                   collection = "C5",
                   subcollection = "GO:BP")
cc_sets <- msigdbr(species = "human",
                   collection = "C5",
                   subcollection = "GO:CC")
# Store all sets together
all_sets <- rbind(hallmark_sets, bp_sets, cc_sets)

# TERM2GENE
term2gene <- all_sets %>%
  select(gs_name, gene_symbol)

# TERM2NAME (readable descriptions)
term2name <- all_sets %>%
  distinct(gs_name, gs_description)

# Background genes
DefaultAssay(SSc_monocytes) <- "RNA"
background_genes <- rownames(SSc_monocytes)
```

Let's run pathway analysis for each direction separately.
 
```{r}
# (1) PBMC

# Add direction (whether expression goes up or down over time) to results data frame
results_sig <- results %>% filter(pval < 0.05) %>% mutate(direction = case_when(
  log2fc > 0 ~ 'UP',
  log2fc < 0 ~ 'DOWN',
))

# Split the dataframe into a list of sub-dataframes: upregulated, downregulated genes
results_list <- split(results_sig, results_sig$direction)

# Run enricher R on each set of genes (up set and down set)
gsea <- lapply(names(results_list),
              function(x) enricher(
                  gene = results_list[[x]]$name,
                  TERM2GENE = term2gene,
                  TERM2NAME = term2name,
                  universe = background_genes,
                  pvalueCutoff = 1,
                  qvalueCutoff = 1,
                  minGSSize = 5,
                  maxGSSize = 500
              )) %>% setNames(names(results_list))

saveRDS(gsea, paste0(project_path, "data/RDS_files/monocyte_gsea.RDS"))


# View results
# DOWN
gsea$DOWN@result %>% head()
# UP
gsea$UP@result %>% head()
```

Now, we save
a) genes that go down
b) genes that go up
c) pathways that go down
d) pathways that go up

```{r}
# # Down genes
down_res <- results_sig %>% filter(log2fc < 0)
write_csv(down_res, paste0(results_dir, "down_DEGs.csv"))
down_res <- read_csv(paste0(results_dir, "down_DEGs.csv"))

# # Up genes
up_res <- results_sig %>% filter(log2fc > 0)
write_csv(up_res, paste0(results_dir, "up_DEGs.csv"))
up_res <- read_csv(paste0(results_dir, "up_DEGs.csv"))

# # Down pathways
down_pathways <- gsea$DOWN@result
write_csv(down_pathways, paste0(results_dir, "down_GSEA.csv"))
down_pathways <- read_csv(paste0(results_dir, "down_GSEA.csv"))

# # Up pathways
up_pathways <- gsea$UP@result
write_csv(up_pathways, paste0(results_dir, "up_GSEA.csv"))
up_pathways <- read_csv(paste0(results_dir, "up_GSEA.csv"))
```


Let's make a bar chart of the decreasing pathways.

```{r}
gsea_down_df <- gsea$DOWN@result %>% select(ID, p.adjust) %>% filter(p.adjust < 0.01)
rownames(gsea_down_df) <- NULL
gsea_down_df$ID <- sub("^[^_]+_", "", gsea_down_df$ID)
gsea_down_df$ID <- factor(gsea_down_df$ID, levels = rev(gsea_down_df$ID))

gsea_down_plot <- ggplot(data = gsea_down_df, 
                         aes(x = -log10(p.adjust),
                             y = ID)) +
  geom_col(fill = "steelblue") +
  theme_bw() +
  ylab("Gene set") +
  xlab(expression("-log"[10]*"p-value")) +
  theme(text = element_text(color = "black", face = "bold")) +
  ggtitle("Gene sets with decreased activity \nin jSSc monocytes over time since ASCT")

# ggsave(paste0(project_path, "figures/gsea_down_plot.png"), gsea_down_plot, width = 8, height = 6)
```

Same for increasing

```{r}
gsea_up_df <- gsea$UP@result %>% select(ID, p.adjust) %>% filter(p.adjust < 0.01)
rownames(gsea_up_df) <- NULL
gsea_up_df$ID <- sub("^[^_]+_", "", gsea_up_df$ID)
gsea_up_df <- gsea_up_df[!duplicated(gsea_up_df$ID), ]
gsea_up_df$ID <- factor(gsea_up_df$ID, levels = rev(gsea_up_df$ID))

gsea_up_plot <- ggplot(data = gsea_up_df, 
                         aes(x = -log10(p.adjust),
                             y = ID)) +
  geom_col(fill = "magenta3") +
  theme_bw() +
  ylab("Gene set") +
  xlab(expression("-log"[10]*"p-value")) +
  theme(text = element_text(color = "black", face = "bold")) +
  ggtitle("Gene sets with decreased activity \nin jSSc monocytes over time since ASCT")

# ggsave(paste0(project_path, "figures/gsea_up_plot.png"), gsea_up_plot, width = 8, height = 6)
```

Let's plot some genes that show up in our favorite (down) pathways

```{r}
favorite_pathways <- c("HALLMARK_TNFA_SIGNALING_VIA_NFKB", "HALLMARK_INFLAMMATORY_RESPONSE", "HALLMARK_TGF_BETA_SIGNALING", "HALLMARK_IL2_STAT5_SIGNALING", "GOBP_EPITHELIAL_CELL_DIFFERENTIATION")

# fav_path_res <- gsea$DOWN %>% filter(ID %in% favorite_pathways)
# 
# # x <- cnetplot(fav_path_res)
# fav_path_res_top <- fav_path_res
# fav_path_res_top@result$geneID <- sapply(
#   fav_path_res@result$geneID,
#   function(x) {
#     genes <- strsplit(x, "/")[[1]]
#     n <- min(length(genes), 15)       # take up to 15 genes, or fewer if less than 15
#     paste(genes[1:n], collapse = "/")
#   }
# )
# 
# cnetplot(fav_path_res_top, circular = TRUE, colorEdge = TRUE)

```

```{r}
# Get all gene sets
gene_sets <- gsea$DOWN@geneSets
# Extract 5 favorite pathways
fav_gene_sets <- gene_sets[favorite_pathways]



# Make a named vector of p-values for quick lookup
pval_lookup <- results$pval
names(pval_lookup) <- results$name

# Order each pathway's genes by p-value
fav_gene_sets_ordered <- lapply(fav_gene_sets, function(genes) {
  # Keep only genes that exist in results
  genes_in_results <- intersect(genes, names(pval_lookup))
  
  # Order by p-value
  genes_in_results[order(pval_lookup[genes_in_results])]
})

topgenes <- lapply(fav_gene_sets_ordered, function(x){head(x, 5)}) %>% unlist() %>% setNames(NULL) %>% unique()

pathway_by_gene_mat <- sapply(topgenes, function(gene) {
  sapply(fav_gene_sets_ordered, function(x) gene %in% x)
})

heat_df <- pathway_by_gene_mat %>%
  as.data.frame() %>%
  rownames_to_column("pathway") %>%
  pivot_longer(-pathway, names_to = "gene", values_to = "present")

heat_df$gene <- factor(heat_df$gene, levels = topgenes)

heat_df$pathway <- sub("^[^_]+_", "", heat_df$pathway)

gsea_plot_with_genes <- 
  ggplot(heat_df, aes(x = gene, y = pathway, fill = present)) +
    geom_tile(color = "white") +
    scale_fill_manual(values = c("FALSE" = "white", "TRUE" = "firebrick")) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid = element_blank(),
      legend.position = "none"
    ) +
    ylab("Gene Set") +
    xlab("Gene") +
    ggtitle("Top DEGs contributing to pathways with decreased \nactivity in jSSc monocytes over time since ASCT")

ggsave(paste0(project_path, "figures/gsea_down_plot_with_genes.png"), gsea_plot_with_genes, width = 8, height = 6, bg = "white")
```

```{r}
gene_plots <- plot_pb_offset(pbo_monocytes, topgenes, groupBy = "subject")

gene_plots_clean <- lapply(seq_along(gene_plots), function(i) {
  p <- gene_plots[[i]]
  nrow <- 6; ncol <- 3
  row <- (i - 1) %/% ncol
  col <- (i - 1) %% ncol
  
  p <- p + theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    #plot.margin = margin(0.25, 0.25, 0.25, 0.25, "pt"),
    plot.title = element_text(size = 30)
  )
  
  if (row < nrow - 1) {
    #p <- p + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
  }
  
  p
})

gene_plots_combined <- wrap_plots(gene_plots_clean, nrow = 6, ncol = 3) +
  plot_layout(guides = "collect") +
  plot_annotation(
    caption = "Months since transplant"
  ) &
  theme(
    legend.position = "right",
    legend.text = element_text(size = 30, face = "plain"),   # fix bold/oversized
    legend.title = element_text(size = 30, face = "plain"),
    legend.key.size = unit(0.8, "lines"),
    plot.caption = element_text(hjust = 0.5, size = 30),
    plot.margin = margin(50, 10, 10, 30, "pt") 
  )

gene_plots_combined <- wrap_elements(full = gene_plots_combined) +
  labs(tag = "Counts per million (log2 scale)") +
  theme(
    plot.tag = element_text(size = 30, angle = 90, hjust = 0.5, vjust = -1),
    plot.tag.position = "left"
  ) + ggtitle("Downward longitudinal trends for key DEGs \nin jSSc monocytes over time since ASCT") + theme(plot.title = element_text(size = 30), face = "bold")

ggsave(paste0(project_path, "figures/gene_plots_combined.png"), 
       gene_plots_combined, width = 14, height = 24, bg = "white")
```

Some plots to show cluster 8 over time

```{r}
SSc_monocytes <- LoadH5Seurat(paste0(project_path, "data/h5seurat/SSc_monocytes.h5seurat"))

healthy_monocytes <- LoadH5Seurat(paste0(project_path, "data/h5seurat/healthy_monocytes.h5seurat"))

merged <- merge(healthy_monocytes, SSc_monocytes)
DefaultAssay(merged) <- "RNA"
merged$group <- factor(merged$group, levels = c("healthy", "baseline", "six", "twelve", "twen4"))

DefaultAssay(SSc_monocytes) <- "RNA"

DimPlot(SSc_monocytes, 
        reduction = "wnn.umap", 
        label = TRUE, 
        split.by = "group", 
        cells.highlight = Cells(subset(SSc_monocytes, seurat_clusters == 8))) +
  scale_color_manual(
    values = c("Unselected" = "grey80", "Group_1" = "red"),
    breaks = "Group_1",
    labels = "Cluster 8"
  ) +
  labs(color = NULL) +
  ggtitle("jSSc monocytes over time since ASCT") +
  labs(caption = str_wrap("SSc monocytes sub-cluster 8 over time since ASCT. RNA markers for this cluster include SERPINE1, EMP1, RGCC, ATF3, CDKN1A, FOSL1, and CCL2. GSEA suggests increased TNF-Alpha signaling, inflammation, IL2/STAT5 signaling, TGF-Beta signaling, and IL6/JAK/STAT3 in this cluster. Cluster 8 is primarily comprised of cells from baseline and 6 months after ASCT, with the population nearly disappearing at 12 and 24 months after ASCT."), width = 120) +
  theme(plot.caption = element_text(size = 9, hjust = 0))

DotPlot(SSc_monocytes, features = c("SERPINE1", "EMP1", "RGCC", "ATF3", "CDKN1A", "FOSL1", "CCL2"), group.by = "group", dot.scale = 20) + coord_flip() + scale_color_gradient2(low = "steelblue", mid = "white", high = "darkred", midpoint = 0) + ggtitle(str_wrap("Expression of jSSc monocyte cluster 8 DEGs over time since ASCT", width = 35)) + ylab("Timepoint") + xlab("Gene") +
  labs(caption = str_wrap("These genes were upregulated in cluster 8 of the jSSc monocytes. Expression within all monocytes (not only cluster 8) is illustrated above.", width = 60)) +
  theme(plot.caption = element_text(size = 9, hjust = 0))


```

