---
title: "6: Generate supporting data values spreadsheets"
output: html_notebook
---

JCI Insight requires an excel file containing supporting data values for all graphs. We generate the file here, using RDS files saved in previous notebooks.

We start by setting the project path and loading packages.

```{r}
# Set project path
project_path <- "/Users/juliaelrod/Desktop/cmu/research/scleroderma/jSSc_ASCT_monocytes_manuscript/code_for_manuscript/"

# Load packages
library(dplyr)
library(openxlsx2)
library(Seurat)
library(SeuratDisk)
library(tidyr)
library(stringr)
library(purrr)
```

We begin by creating the workbook.

```{r}
wb <- wb_workbook()

# Helper: add a bold title/label cell
wb_add_title <- function(wb, sheet, x, start_row, start_col = 1) {
  wb %>%
    wb_add_data(sheet = sheet, x = x, start_row = start_row, start_col = start_col) %>%
    wb_add_font(sheet = sheet,
                dims = wb_dims(rows = start_row, cols = start_col),
                bold = "1")
}

# Helper: add a data frame with bolded headers and (optionally) bolded first column.
# Numeric columns are rounded to 3 significant figures.
wb_add_table <- function(wb, sheet, x, start_row = 1, start_col = 1, bold_first_col = TRUE) {
  # Round numeric columns to 3 significant figures (base R, tolerates blank colnames)
  num_cols <- vapply(x, is.numeric, logical(1))
  x[num_cols] <- lapply(x[num_cols], signif, 3)
  end_col <- start_col + ncol(x) - 1
  wb <- wb %>%
    wb_add_data(sheet = sheet, x = x, start_row = start_row, start_col = start_col) %>%
    wb_add_font(sheet = sheet,
                dims = wb_dims(rows = start_row, cols = start_col:end_col),
                bold = "1")
  if (bold_first_col && nrow(x) > 0) {
    wb <- wb %>%
      wb_add_font(sheet = sheet,
                  dims = wb_dims(rows = (start_row + 1):(start_row + nrow(x)),
                                 cols = start_col),
                  bold = "1")
  }
  wb
}
```

1. Figure 2F

```{r}
dat <- readRDS(paste0(project_path, "data/supporting_data_values/2F.RDS"))

# Set sheet name
sheet_name <- paste0("Fig. 2F")

# Add to workbook
wb <- wb %>%
  wb_add_worksheet(sheet_name) %>%
  wb_add_table(sheet = sheet_name, x = dat)
```

2. Figure 3A

```{r}
dat <- readRDS(paste0(project_path, "data/supporting_data_values/3A.RDS"))

# Set sheet name
sheet_name <- paste0("Fig. 3A")

# Add to workbook
wb <- wb %>%
  wb_add_worksheet(sheet_name) %>%
  wb_add_table(sheet = sheet_name, x = dat)
```

3. Figure 3B

```{r}
dat <- readRDS(paste0(project_path, "data/supporting_data_values/3B.RDS"))

# Set sheet name
sheet_name <- paste0("Fig. 3B")

# Add to workbook
wb <- wb %>%
  wb_add_worksheet(sheet_name) %>%
  wb_add_table(sheet = sheet_name, x = dat)
```

4. Supplemental figure 3A

```{r}
dat <- readRDS(paste0(project_path, "data/supporting_data_values/Supp3A.RDS"))

# Set sheet name
sheet_name <- paste0("S Fig. 3A")

# Add to workbook
wb <- wb %>%
  wb_add_worksheet(sheet_name) %>%
  wb_add_table(sheet = sheet_name, x = dat)
```

5. Supplemental figure 3B

```{r}
dat <- readRDS(paste0(project_path, "data/supporting_data_values/Supp3B.RDS"))

# Set sheet name
sheet_name <- paste0("S Fig. 3B")

# Add to workbook
wb <- wb %>%
  wb_add_worksheet(sheet_name) %>%
  wb_add_table(sheet = sheet_name, x = dat)
```

6. Supplemental figure 3C

```{r}
dat <- readRDS(paste0(project_path, "data/supporting_data_values/Supp3C.RDS"))

# Set sheet name
sheet_name <- paste0("S Fig. 3C")

# Add to workbook
wb <- wb %>%
  wb_add_worksheet(sheet_name) %>%
  wb_add_table(sheet = sheet_name, x = dat)
```

7. Supplemental figure 4

```{r}
dat <- readRDS(paste0(project_path, "data/supporting_data_values/Supp4.RDS"))

# Set sheet name
sheet_name <- paste0("S Fig. 4")

# Add to workbook
wb <- wb %>%
  wb_add_worksheet(sheet_name) %>%
  wb_add_table(sheet = sheet_name, x = dat)
```

8. Supplemental figure 5

```{r}
dat <- readRDS(paste0(project_path, "data/supporting_data_values/Supp5.RDS"))

# Set sheet name
sheet_name <- paste0("S Fig. 5")

# Add to workbook
wb <- wb %>%
  wb_add_worksheet(sheet_name) %>%
  wb_add_table(sheet = sheet_name, x = dat)
```

9. Supplemental figure 6A

```{r}
dat <- readRDS(paste0(project_path, "data/supporting_data_values/Supp6A.RDS"))

# Set sheet name
sheet_name <- paste0("S Fig. 6A")

# Add to workbook
wb <- wb %>%
  wb_add_worksheet(sheet_name) %>%
  wb_add_table(sheet = sheet_name, x = dat)
```

10. Supplemental figure 6B

```{r}
dat <- readRDS(paste0(project_path, "data/supporting_data_values/Supp6B.RDS"))

# Set sheet name
sheet_name <- paste0("S Fig. 6B")

# Add to workbook
wb <- wb %>%
  wb_add_worksheet(sheet_name) %>%
  wb_add_table(sheet = sheet_name, x = dat)
```

Format and save the workbook

```{r}
# Auto-fit column widths on every sheet.
for (sn in wb$get_sheet_names()) {
  df <- wb_to_df(wb, sheet = sn, col_names = FALSE)
  if (is.null(df) || ncol(df) == 0) next

  widths <- vapply(df, function(col) {
    vals <- as.character(col)
    vals <- vals[!is.na(vals) & vals != ""]
    if (length(vals) == 0) return(8.43)              # Excel default width
    min(max(nchar(vals)) + 2, 255)                   # +2 padding, cap at Excel max
  }, numeric(1))

  wb <- wb_set_col_widths(wb, sheet = sn, cols = seq_along(widths), widths = widths)
}

wb_save(wb, paste0(project_path, "data/supporting_data_values/supporting_data_values.xlsx"), overwrite = T)
```

