---
title: "jSSc ASCT monocytes longitudinal analysis"
output: html_notebook
---

In this notebook, we run our longitudinal monocyte pseudobulk offset analysis. We use a single model which includes an indicator term for tissue. 

We begin by loading in useful packages and custom functions.

First, clear the environment and set the path where the project folder is saved.

```{r}
# Clear all objects from environment
rm(list = ls(all.names = TRUE))
# Free up memory
gc()
# Increase max vector memory
mem.maxVSize(200000)

project_path <- "/Users/juliaelrod/Desktop/cmu/research/scleroderma/jSSc_ASCT_monocytes_manuscript/code_for_manuscript/"

results_dir <- paste0(project_path, "data/DEG_pathway_CSVs/")
```

Now load in useful packages and custom functions.

```{r}
# Load packages
packages <- c("Seurat", "dplyr", "tidyr", "SeuratDisk", "harmony", "readr",
              "ggplot2", "stringr", "SoupX", "dsb", "data.table", "Matrix",
              "scDblFinder", "scater", "ggrepel", "tidyverse", "RColorBrewer", "pheatmap", "clusterProfiler", "enrichplot", "msigdbr")
invisible(lapply(packages, require, character.only = TRUE))

# Source functions
function_files <- list.files(paste0(project_path, "functions"), pattern = "\\.R$", full.names = TRUE)
invisible(lapply(function_files, source))

# Function to update cell_types
update_cell_types <- function(df, cluster, cell_type = NA) {
  df[df$seurat_clusters == cluster, "cell_type"] <- cell_type
  return(df)
}
```
We load in the monocytes and cluster them. 

```{r}
# # scler_pbmc <- LoadH5Seurat(paste0(project_path, "data/h5seurat/scler_pbmc_10.h5seurat"))
# # monocytes <- subset(scler_pbmc, cell_type == "monocyte")
# # 
# # monocytes <- sub_cluster(monocytes,
# #                          res = 0.8)
# # SaveH5Seurat(monocytes, paste0(project_path, "data/h5seurat/monocytes.h5seurat"))
# 
# # Load in monocytes
# monocytes <- LoadH5Seurat(paste0(project_path, "data/h5seurat/monocytes.h5seurat"))
# # Set default assay
# DefaultAssay(monocytes) <- "RNA"
# 
# # Look at group proportions for each cluster
# all_cluster_prop <- 
# monocytes$seurat_clusters %>% levels() %>% lapply(
#   function(clus){
#     (monocytes@meta.data %>% filter(seurat_clusters == clus) %>% pull(group) %>% table()) / (monocytes@meta.data %>% pull(group) %>% table())
#   }
# ) %>% setNames(paste0("cluster_", (monocytes$seurat_clusters %>% levels())))
# 
# # Pull out proportions for cluster 9 and plot
# clus9_pct_df <- as.data.frame(all_cluster_prop$cluster_9 * 100) 
# # Rename columns
# colnames(clus9_pct_df) <- c("Group", "Pct")
# # Rename groups
# clus9_pct_df <- clus9_pct_df %>% mutate(
#   Group = factor(case_match(
#     Group,
#     "healthy" ~ "Healthy",
#     "baseline" ~ "SSc baseline",
#     "six" ~ "SSc 6 months",
#     "twelve" ~ "SSc 12 months",
#     "twen4" ~ "SSc 24 months"
#   ), levels = c("Healthy", "SSc baseline", "SSc 6 months", "SSc 12 months", "SSc 24 months"))
# )
# # Plot pct in each group comprised by cluster 9
# clus9_pct_plot <- ggplot(aes(x = Group, y = Pct), data = clus9_pct_df) +
#   geom_col(fill = c("red", rep("blue", 4))) + ylab("Percentage (%)") + xlab("") + theme_bw() + ggtitle(str_wrap("Percentage of monocytes in each group compromised by cluster 9", width = 45))
# # Save image
# ggsave(paste0(project_path, "march_15_analysis/clus9_pct_plot.png"),
#        clus9_pct_plot,
#        width = 6,
#        height = 7)
# 
# # Get RNA markers for each cluster
# rna_markers <- FindAllMarkers(monocytes, assay = "RNA", only.pos = T)
# monocytes$seurat_clusters %>% levels() %>% lapply(function(clus){
#   rna_markers %>% filter(cluster == clus) %>% head()
# }) %>% setNames(paste0("cluster_", (monocytes$seurat_clusters %>% levels())))
# 
# # Ok, cluster 9 in the combined HC/jSSc monocytes is equivalent to cluster 8 in the jSSc monocytes alone.
# rna_markers %>% filter(cluster == 9 & gene %in% c("SERPINE1", "EMP1", "RGCC", "ATF3", "CDKN1A", "FOSL1", "CCL2"))
# # Save markers for cluster 9
# rna_markers %>% filter(cluster == 9) %>% write_csv(paste0(project_path, "march_15_analysis/clus9_markers.csv"))
```

Let's run pathway analysis on the markers from cluster 9. 

```{r}
# # Get hallmark, GO biological process, and GO cellular component pathways
# hallmark_sets <- msigdbr(species = "human",
#                          collection = "H")
# bp_sets <- msigdbr(species = "human",
#                    collection = "C5",
#                    subcollection = "GO:BP")
# cc_sets <- msigdbr(species = "human",
#                    collection = "C5",
#                    subcollection = "GO:CC")
# # Store all sets together
# all_sets <- rbind(hallmark_sets, bp_sets, cc_sets)
# 
# # TERM2GENE
# term2gene <- all_sets %>%
#   select(gs_name, gene_symbol)
# 
# # TERM2NAME (readable descriptions)
# term2name <- all_sets %>%
#   distinct(gs_name, gs_description)
# 
# # Background genes
# background_genes <- rownames(monocytes)
# 
# # Run gsea
# clus9_degs <- rna_markers %>% filter(cluster == 9) %>% pull(gene)
# clus9_gsea <- enricher(gene = clus9_degs,
#                   TERM2GENE = term2gene,
#                   TERM2NAME = term2name,
#                   universe = background_genes,
#                   minGSSize = 5,
#                   maxGSSize = 500,
#               )
# # Filter to only included adjusted p value < 0.05
# clus9_pathways <- clus9_gsea@result %>% filter(p.adjust < 0.05)
# # Save as CSV
# clus9_pathways %>% write_csv(paste0(project_path, "march_15_analysis/clus9_pathways.csv"))
```

We will also compare SSc and healthy within cluster 9. 

```{r}
# # Set ident to health
# Idents(monocytes) <- "health"
# # Compare SSc and healthy in cluster 9
# SSc_v_healthy_clus_9_markers <- FindMarkers(subset(monocytes, seurat_clusters == 9), ident.1 = "SSc", ident.2 = "healthy") 
# # Save
# SSc_v_healthy_clus_9_markers %>% write_csv(paste0(project_path, "march_15_analysis/clus9_SSc_v_healthy_markers.csv"))
# # Set ident back to Seurat clusters
# Idents(monocytes) <- "seurat_clusters"
```

Now, we'll make a UMAP comparing this cluster across groups, and a dotplot comparing the gene expression. 

```{r}
# clus9_umap_by_group <- 
# DimPlot(monocytes, 
#         reduction = "wnn.umap", 
#         label = TRUE, 
#         split.by = "group", 
#         cells.highlight = Cells(subset(monocytes, seurat_clusters == 9))) +
#   scale_color_manual(
#     values = c("Unselected" = "grey80", "Group_1" = "red"),
#     breaks = "Group_1",
#     labels = "Cluster 9"
#   ) +
#   labs(color = NULL) +
#   ggtitle("Healthy control and jSSc monocytes over \ntime since ASCT") +
#   labs(caption = str_wrap("SSc and healthy monocytes sub-cluster 9 over time since ASCT. RNA markers for this cluster include SERPINE1, EMP1, RGCC, ATF3, CDKN1A, FOSL1, and CCL2. GSEA suggests increased TNF-Alpha signaling, inflammation, IL2/STAT5 signaling, TGF-Beta signaling, and epithelial cell differentiation in this cluster. Cluster 9 is primarily comprised of healthy cells and cells from baseline and 6 months after ASCT, with the population nearly disappearing in patients at 12 and 24 months after ASCT."), width = 120) +
#   theme(plot.caption = element_text(size = 9, hjust = 0))
# ggsave(paste0(project_path, "march_15_analysis/clus9_umap_by_group.png"), clus9_umap_by_group, width = 7, height = 6)
```

Some plots to show cluster 9 over time

```{r}
# dotplot_clus9_degs <- 
# DotPlot(monocytes, 
#         features = c("SERPINE1", "EMP1", "RGCC", "ATF3", "CDKN1A", "FOSL1", "CCL2"), 
#         group.by = "group", 
#         dot.scale = 20) + 
#   coord_flip() + 
#   scale_color_gradient2(low = "steelblue", mid = "white", high = "darkred", midpoint = 0) + 
#   ggtitle(str_wrap("Expression of jSSc monocyte cluster 9 DEGs over time since ASCT", width = 35)) + 
#   ylab("Timepoint") + 
#   xlab("Gene") +
#   labs(caption = str_wrap("These genes were upregulated in cluster 9 of the jSSc monocytes. Expression within all monocytes (not only cluster 9) is illustrated above.", width = 60)) +
#   theme(plot.caption = element_text(size = 9, hjust = 0))
# 
# # Save image
# ggsave(paste0(project_path, "march_15_analysis/clus9_deg_dotplot.png"), dotplot_clus9_degs, width = 7, height = 6, bg = "white")
```

Now we're going to annotate the monocytes (with all HC and SSc cells together). In a separate notebook, we did this separately for HCs and for each patient.

```{r}
# cell_types <- data.frame(seurat_clusters = levels(monocytes$seurat_clusters),
#                           cell_type = NA)
# 
# monocytes %>% DimPlot(reduction = "wnn.umap", label = T)
# 
# adt_markers <- FindAllMarkers(monocytes, assay = "DSB", only.pos = T)
# 
# monocytes$seurat_clusters %>% levels() %>% lapply(function(clus){
#   rna_markers %>% filter(cluster == clus) %>% head() %>% rbind((
#   adt_markers %>% filter(cluster == clus) %>% head()))
# }) %>% setNames(paste0("cluster_", (monocytes$seurat_clusters %>% levels())))
# 
# FeaturePlot(monocytes, "CD14", label = T, reduction = "wnn.umap")
# for (clus in c(0, 1, 2, 3, 4, 7, 9)) {
#   rna_markers %>% filter(cluster == clus & gene == "CD14") %>% rbind((
#   adt_markers %>% filter(cluster == clus & gene == "anti-human-CD14"))) %>% print()
#   cell_types <- update_cell_types(cell_types, clus, cell_type = "CD14+")
# }
# 
# FeaturePlot(monocytes, "FCGR3A", label = T, reduction = "wnn.umap")
# for (clus in c(5, 6)) {
#   rna_markers %>% filter(cluster == clus & gene == "FCGR3A") %>% rbind((
#   adt_markers %>% filter(cluster == clus & gene == "anti-human-CD16"))) %>% print()
#   cell_types <- update_cell_types(cell_types, clus, cell_type = "CD16+")
# }
# 
# # Now, we just need to figure out clusters 8, 10, and 11
# c(8, 10, 11) %>% lapply(function(clus){
#   rna_markers %>% filter(cluster == clus) %>% head() %>% rbind((
#   adt_markers %>% filter(cluster == clus) %>% head()))
# }) %>% setNames(paste0("cluster_", c(8, 10, 11)))
# 
# cell_types <- update_cell_types(cell_types, 8, "B-like")
# cell_types <- update_cell_types(cell_types, 10, "T-like")
# cell_types <- update_cell_types(cell_types, 11, "NK-like")
# 
# # Add cell types to Seurat object
# cell_type_vec <-
#    setNames(as.character(cell_types$cell_type),
#           as.character(cell_types$seurat_clusters))[as.character(monocytes$seurat_clusters)]
# names(cell_type_vec) <- NULL
# monocytes$cell_type <- cell_type_vec
# 
# Plot it
monocyte_dim_plot <-
  monocytes %>% DimPlot(reduction = "wnn.umap", label = T, group.by = "cell_type", cols = brewer.pal(8, "Set2")) + ggtitle("jSSc and healthy monocytes")
# Save
ggsave(paste0(project_path, "figures/jSSc_and_healthy_monocytes_umap.png"))

monocyte_dim_plot_by_group <-
  monocytes %>% DimPlot(reduction = "wnn.umap", label = T, group.by = "cell_type", cols = brewer.pal(8, "Set2"), split.by = "group") + ggtitle("jSSc and healthy monocytes by group")
ggsave(paste0(project_path, "figures/jSSc_and_healthy_monocytes_by_group_umap.png"), height = 4, width = 9)

monocyte_dim_plot_by_health <-
  monocytes %>% DimPlot(reduction = "wnn.umap", label = T, group.by = "cell_type", cols = brewer.pal(8, "Set2"), split.by = "health") + ggtitle("jSSc vs. healthy monocytes")
ggsave(paste0(project_path, "figures/jSSc_and_healthy_monocytes_by_health_umap.png"), height = 5.79, width = 7.4)

monocyte_dim_plot_by_subject <-
  monocytes %>% DimPlot(reduction = "wnn.umap", label = T, group.by = "cell_type", cols = brewer.pal(8, "Set2"), split.by = "subject") + ggtitle("jSSc and healthy monocytes by subject")
ggsave(paste0(project_path, "figures/jSSc_and_healthy_monocytes_by_subject_umap.png"), height = 3, width = 15)
# 
# # Save annotated Seurat object
# SaveH5Seurat(monocytes, paste0(project_path, "data/h5seurat/monocytes_2.h5seurat"))
```

Now, we make heatmaps to illustrate the markers used for the above annotation.

```{r}
# Load in annotated monocytes
monocytes <- LoadH5Seurat(paste0(project_path, "data/h5seurat/monocytes_2.h5seurat"))

# FindAllMarkers by cell type
Idents(monocytes) <- "cell_type"
cell_type_markers_RNA <- FindAllMarkers(monocytes, assay = "RNA", only.pos = T)
cell_type_markers_ADT <- FindAllMarkers(monocytes, assay = "DSB", only.pos = T)

# CD14 markers
CD14_rna_markers <- c("CD14")
CD14_adt_markers <- c("anti-human-CD14")

# CD16 markers
CD16_rna_markers <- c("FCGR3A")
CD16_adt_markers <- c("anti-human-CD16")

# B-like markers
B_like_rna_markers <- c("CD79A", "MS4A1")
B_like_adt_markers <- c("anti-human-CD19", "anti-human-IgD")

# T-like markers
T_like_rna_markers <- c("CD3E", "CD3D")
T_like_adt_markers <- c("anti-human-CD3")

# NK-like markers
NK_like_rna_markers <- c("NKG7", "KLRD1")
NK_like_adt_markers <- c("anti-human-CD56")

# Put all RNA markers together
rna_cell_type_markers <- c(CD14_rna_markers,
                 CD16_rna_markers,
                 B_like_rna_markers,
                 T_like_rna_markers,
                 NK_like_rna_markers)

# Put all ADT markers together
adt_cell_type_markers <- c(CD14_adt_markers,
                 CD16_adt_markers,
                 B_like_adt_markers,
                 T_like_adt_markers,
                 NK_like_adt_markers)

# Order cell types to match order of markers
monocytes$cell_type <- factor(monocytes$cell_type,
                              levels = unique(monocytes$cell_type))

rna_marker_heatmap <-
    DoHeatmap(subset(monocytes, downsample = 60),
              features = rna_cell_type_markers,
              assay = "RNA",
              group.bar.height = 0,
              size = 5,
              lines.width = 3) +
    scale_fill_gradient(low = "grey95", high = "darkblue", na.value = "white") +
    theme(legend.position = "right",
          axis.text.y = element_text(size = 15),
          plot.title = element_text(size = 25, face = "bold")) +
    guides(color = "none") +
  ggtitle("RNA markers")
ggsave(paste0(project_path, "figures/monocytes_rna_marker_heatmap.png"), rna_marker_heatmap, width = 8, height = 6)

label_map <- setNames(gsub("^anti-human-", "", adt_cell_type_markers), adt_cell_type_markers)
adt_marker_heatmap <-
    DoHeatmap(subset(monocytes, downsample = 60),
              features = adt_cell_type_markers,
              assay = "DSB",
              group.bar.height = 0,
              size = 5,
              lines.width = 8,
              disp.min = -0.5,
              disp.max = 1) +
    scale_fill_gradient(low = "grey95", high = "darkred", na.value = "black") +
      scale_y_discrete(labels = label_map) +
    ggtitle("ADT markers") +
    theme(legend.position = "right",
          axis.text.y = element_text(size = 15),
          plot.title = element_text(size = 25, face = "bold")) +
  guides(color = "none")
ggsave(paste0(project_path, "figures/monocytes_adt_marker_heatmap.png"), adt_marker_heatmap, width = 8, height = 6)

rna_marker_dotplot <-
  DotPlot(monocytes,
          features = rna_cell_type_markers,
          group.by = "cell_type",
          assay = "RNA") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
ggsave(paste0(project_path, "figures/monocytes_rna_marker_dotplot.png"), rna_marker_dotplot, width = 6, height = 6)

adt_marker_dotplot <-
  DotPlot(monocytes,
          features = adt_cell_type_markers,
          group.by = "cell_type",
          assay = "DSB",
          cols = c("lightgrey", "darkred")) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
ggsave(paste0(project_path, "figures/monocytes_adt_marker_dotplot.png"), adt_marker_dotplot, width = 6, height = 6)

# Cell type proportions plot
# Add a blank level between HC and SSc baseline
monocytes@meta.data$group_spaced <- factor(
  monocytes@meta.data$group,
  levels = c(levels(monocytes@meta.data$group)[1], "", 
             levels(monocytes@meta.data$group)[2:5])
)

monocytes_prop_by_group <- ggplot(monocytes@meta.data, aes(y = group_spaced, fill = cell_type)) +
    geom_bar(position = "fill", color = "black") +
    scale_fill_brewer(palette = "Set3", name = "Cell type") +
    scale_y_discrete(labels = c("HC", "", "SSc baseline", "SSc 6 mo.", "SSc 12 mo.", "SSc 24 mo."),
                     drop = FALSE) +
    coord_flip() +
    theme_bw() +
    labs(x = "Cell type proportion", y = "Group") +
    geom_hline(yintercept = 2, col = "black", linewidth = 1, linetype = "dashed") +
    ggtitle(str_wrap("HC vs. jSSc monocyte subtype proportions over time since ASCT", width = 40)) +
    theme(
        axis.text.x = element_text(size = 14, angle = 45, hjust = 1),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_text(size = 16),
        axis.title.y = element_text(size = 16),
        plot.title = element_text(size = 18),
        legend.text = element_text(size = 14)
    )
ggsave(paste0(project_path, "figures/monocytes_prop_by_group.png"), monocytes_prop_by_group, height = 5.76, width = 7)
```

Now, we will make a volcano plot comparing SSc and HC in the monocytes.

```{r}
# DefaultAssay(monocytes) <- "RNA"
# 
# SSc_v_h_PB <- pb_offset(subset(monocytes, group %in% c("baseline", "healthy")), "sample_id", ~health)
# SSc_v_h_PB_results <- SSc_v_h_PB$fit %>% test_de("healthSSc") %>% arrange(pval)
# 
# monocytes$cell_id <- colnames(monocytes)
# SSc_v_h_SC <- pb_offset(subset(monocytes, group %in% c("baseline", "healthy")), "cell_id", ~health)
# SSc_v_h_SC_results <- SSc_v_h_SC$fit %>% test_de("healthSSc") %>% arrange(pval)
# 
# top10_up_pb <- SSc_v_h_PB_results %>% arrange(pval) %>% filter(lfc > 0.6) %>% pull(name) %>% head(10)
# 
# top10_down_pb <- SSc_v_h_PB_results %>% arrange(pval) %>% filter(lfc < -0.6) %>% pull(name) %>% head(10)
# 
# SSc_v_h_PB_results <- SSc_v_h_PB_results %>%
#   dplyr::rename(log2fc = lfc) %>%
#   mutate(`SSc vs. HC` = case_when(
#     log2fc > 0.6 & pval < 0.05 ~ "Up",
#     log2fc < -0.6 & pval < 0.05 ~ "Down",
#     TRUE ~ "Not significant"
#   )) %>%
#   mutate(delabel = case_when(
#     name %in% c(top10_down_pb, top10_up_pb) ~ name,
#     TRUE ~ NA
#   ))
# 
# SSc_v_h_PB_results$`SSc vs. HC` <- factor(SSc_v_h_PB_results$`SSc vs. HC`, levels = c("Up", "Down", "Not significant"))
# 
# volcanoplot_pb <- ggplot(data = SSc_v_h_PB_results, aes(x = log2fc, y = -log10(pval), col = `SSc vs. HC`, label = delabel)) +
#   geom_vline(xintercept = c(-0.6, 0.6), col = "gray", linetype = 'dashed') +
#   geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed') + 
#   geom_point(size = 2) +
#   theme_classic(base_size = 20) +
#   theme(axis.title.y = element_text(face = "bold", margin = margin(0,20,0,0), size = rel(1.1), color = 'black'),
#         axis.title.x = element_text(hjust = 0.5, face = "bold", margin = margin(20,0,0,0), size = rel(1.1), color = 'black'),
#         plot.title = element_text(hjust = 0.5, size = rel(1.1))) +
#   scale_color_manual(values = c( "red3", "#00AFBB", "grey")) +
#   labs(x = expression("log"[2]*"FC"), y = expression("-log"[10]*"p-value")) + 
#   scale_x_continuous(breaks = seq(-10, 10, 2)) +
#   ggtitle(str_wrap("Differentially expressed genes in SSc (before ASCT) vs. healthy controls (pseudobulk level)", width = 50)) +
#   geom_label_repel(max.overlaps = Inf,
#                   size = rel(3),
#                   box.padding = 0.3,
#                   force = 4,
#                   show.legend = F)
# ggsave(paste0(project_path, "figures/SSc_v_healthy_pb_volcano.png"), width = 8, height = 6)
# 
# top10_up_sc <- SSc_v_h_SC_results %>% arrange(pval) %>% filter(lfc > 0.6) %>% pull(name) %>% head(10)
# 
# top10_down_sc <- SSc_v_h_SC_results %>% arrange(pval) %>% filter(lfc < -0.6) %>% pull(name) %>% head(10)
# 
# SSc_v_h_SC_results <- SSc_v_h_SC_results %>%
#   dplyr::rename(log2fc = lfc) %>%
#   mutate(`SSc vs. HC` = case_when(
#     log2fc > 0.6 & pval < 0.05 ~ "Up",
#     log2fc < -0.6 & pval < 0.05 ~ "Down",
#     TRUE ~ "Not significant"
#   )) %>%
#   mutate(delabel = case_when(
#     name %in% c(top10_down_sc, top10_up_sc) ~ name,
#     TRUE ~ NA
#   ))
# 
# volcanoplot_sc <- ggplot(data = SSc_v_h_SC_results, aes(x = log2fc, y = -log10(pval), col = `SSc vs. HC`, label = delabel)) +
#   geom_vline(xintercept = c(-0.6, 0.6), col = "gray", linetype = 'dashed') +
#   geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed') + 
#   geom_point(size = 2) +
#   theme_classic(base_size = 20) +
#   theme(axis.title.y = element_text(face = "bold", margin = margin(0,20,0,0), size = rel(1.1), color = 'black'),
#         axis.title.x = element_text(hjust = 0.5, face = "bold", margin = margin(20,0,0,0), size = rel(1.1), color = 'black'),
#         plot.title = element_text(hjust = 0.5, size = rel(1.1))) +
#   scale_color_manual(values = c("#00AFBB", "grey", "red3")) +
#   labs(x = expression("log"[2]*"FC"), y = expression("-log"[10]*"p-value")) + 
#   scale_x_continuous(breaks = seq(-10, 10, 2)) +
#   ggtitle("Differentially expressed genes in SSc (before ASCT) vs. healthy controls (single cell level)") +
#   geom_label_repel(max.overlaps = Inf,
#                   size = rel(3),
#                   box.padding = 0.3,
#                   force = 4,
#                   show.legend = F)
```

Ok, now I will make a new version to include of the SRF grant 

It will include the following genes in monocytes:
SERPINE1, FOS1, SMAD3, TAGLN2, CDKNiC

And I will show
a. Dotplots of expression over time since ASCT compared to healthy controls
b. Line plots showing the longitudinal trend over time 
c. Tile plot showing which key pathways these genes show up in

```{r}
srf_genes <- c("SERPINE1", "FOSL1", "SMAD3", "TAGLN2", "CDKN1C")
srf_dotplot <-
DotPlot(monocytes,
        features = srf_genes,
        group.by = "group",
        dot.scale = 30) +
  coord_flip() +
  scale_color_gradient2(low = "steelblue", mid = "white", high = "darkred", midpoint = 0) +
  ggtitle(str_wrap("Expression of jSSc monocyte DEGs over time since ASCT", width = 70)) +
  ylab("Timepoint") +
  xlab("Gene") +
  theme(plot.caption = element_text(size = 9, hjust = 0))

# Save image
ggsave(paste0(project_path, "march_23_analysis/srf_dotplot.png"), srf_dotplot, width = 10, height = 6, bg = "white")
```

```{r}
# Design
design_formula <- ~ subject + poly(timepoint)

# Fit offset pseudobulk model
pbo_monocytes <- pb_offset(
  seurat_obj = subset(monocytes, health == "SSc"),
  sample_id_variable = "sample_id",
  design_formula = design_formula)

gene_plots <- plot_pb_offset(pbo_monocytes, srf_genes, groupBy = "subject")

gene_plots_clean <- lapply(seq_along(gene_plots), function(i) {
  p <- gene_plots[[i]]
  nrow <- 6; ncol <- 3
  row <- (i - 1) %/% ncol
  col <- (i - 1) %% ncol
  
  p <- p + theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    #plot.margin = margin(0.25, 0.25, 0.25, 0.25, "pt"),
    plot.title = element_text(size = 30)
  )
  
  if (row < nrow - 1) {
    #p <- p + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
  }
  
  p
})

gene_plots_combined <- wrap_plots(gene_plots_clean, nrow = 2, ncol = 3) +
  plot_layout(guides = "collect") +
  plot_annotation(
    caption = "Months since transplant"
  ) &
  theme(
    legend.position = "right",
    legend.text = element_text(size = 30, face = "plain"),   # fix bold/oversized
    legend.title = element_text(size = 30, face = "plain"),
    legend.key.size = unit(0.8, "lines"),
    plot.caption = element_text(hjust = 0.5, size = 30),
    plot.margin = margin(50, 10, 10, 30, "pt") 
  )

gene_plots_combined <- wrap_elements(full = gene_plots_combined) +
  labs(tag = "Counts per million (log2 scale)") +
  theme(
    plot.tag = element_text(size = 30, angle = 90, hjust = 0.5, vjust = -1),
    plot.tag.position = "left"
  ) + ggtitle("Downward longitudinal trends for key DEGs in jSSc monocytes over \ntime since ASCT") + theme(plot.title = element_text(size = 30), face = "bold")

ggsave(paste0(project_path, "march_23_analysis/gene_plots_combined.png"), 
       gene_plots_combined, width = 14, height = 8, bg = "white")
```

```{r}
# Load GSEA
gsea <- readRDS(paste0(project_path, "data/RDS_files/monocyte_gsea.RDS"))

favorite_pathways <- c("HALLMARK_TNFA_SIGNALING_VIA_NFKB", "HALLMARK_INFLAMMATORY_RESPONSE", "HALLMARK_TGF_BETA_SIGNALING", "HALLMARK_IL2_STAT5_SIGNALING", "GOBP_EPITHELIAL_CELL_DIFFERENTIATION")

# Get all gene sets
gene_sets <- gsea$DOWN@geneSets
# Extract 5 favorite pathways
fav_gene_sets <- gene_sets[favorite_pathways]

# Make a named vector of p-values for quick lookup
results <- test_de(pbo_monocytes$fit, `poly(timepoint)`) %>% 
  arrange(pval)
pval_lookup <- results$pval
names(pval_lookup) <- results$name

# Order each pathway's genes by p-value
fav_gene_sets_ordered <- lapply(fav_gene_sets, function(genes) {
  # Keep only genes that exist in results
  genes_in_results <- intersect(genes, names(pval_lookup))
  
  # Order by p-value
  genes_in_results[order(pval_lookup[genes_in_results])]
})

pathway_by_gene_mat <- sapply(srf_genes, function(gene) {
  sapply(fav_gene_sets_ordered, function(x) gene %in% x)
})

heat_df <- pathway_by_gene_mat %>%
  as.data.frame() %>%
  rownames_to_column("pathway") %>%
  pivot_longer(-pathway, names_to = "gene", values_to = "present")

heat_df$gene <- factor(heat_df$gene, levels = srf_genes)

heat_df$pathway <- sub("^[^_]+_", "", heat_df$pathway)

gsea_plot_with_genes <- 
  ggplot(heat_df, aes(x = gene, y = pathway, fill = present)) +
    geom_tile(color = "white") +
    scale_fill_manual(values = c("FALSE" = "white", "TRUE" = "firebrick")) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      text = element_text(color = "black", face = "bold", size = 15),
      panel.grid = element_blank(),
      legend.position = "none",
      plot.title = element_text(hjust = 0),
      plot.title.position = "plot"
    ) +
    ylab("Gene Set") +
    xlab("Gene") +
    ggtitle(str_wrap("Pathways and DEGs with decreased activity in jSSc monocytes over time since ASCT", width = 60))

ggsave(paste0(project_path, "march_23_analysis/gsea_down_plot_with_genes.png"), gsea_plot_with_genes, width = 8, height = 6, bg = "white")



```


```{r}
# Pathways selected by Anwesha
down_pathways_to_show <- data.frame(full_name = c("HALLMARK_TNFA_SIGNALING_VIA_NFKB",
                           "HALLMARK_INFLAMMATORY_RESPONSE",
                           "GOBP_EPITHELIAL_CELL_DIFFERENTIATION",
                           "HALLMARK_TGF_BETA_SIGNALING",
                           "GOBP_REGULATION_OF_EPITHELIAL_CELL_PROLIFERATION",
                           "GOBP_EPITHELIAL_CELL_PROLIFERATION",
                           "HALLMARK_IL2_STAT5_SIGNALING",
                           "HALLMARK_APOPTOSIS"
                           ), plot_name = c("TNF-α signaling through NF-κB (Hallmark)",
                                            "Inflammatory response (Hallmark)",
                                            "Epithelial cell differentiation (GOBP)",
                                            "TGF-β signaling (Hallmark)",
                                            "Regulation of epithelial cell proliferation (GOBP)",
                                            "Epithelial cell proliferation (GOBP)",
                                            "IL2-STAT-5 signaling (Hallmark)",
                                            "Apoptosis (Hallmark)"))

gsea_down_df <- down_pathways_to_show %>%
  inner_join(
    gsea$DOWN@result %>% select(ID, p.adjust),
    by = c("full_name" = "ID")
  ) %>%
  filter(p.adjust < 0.05)

gsea_down_plot <- ggplot(data = gsea_down_df, 
                         aes(x = -log10(p.adjust),
                             y = reorder(plot_name, -p.adjust))) +
  geom_col(fill = "steelblue") +
  theme_bw() +
  ylab("Gene set") +
  xlab(expression("-log"[10]*"p-value")) +
  ggtitle("Gene sets with decreased activity \nin jSSc monocytes over time since ASCT") +
  theme(
    text = element_text(face = "bold"),
    axis.text = element_text(color = "black", size = 13),
    axis.title = element_text(color = "black", size = 15),
    plot.title = element_text(color = "black", size = 15)
  ) +
  scale_y_discrete(labels = function(x) str_wrap(x, width = 25))

ggsave(paste0(project_path, "figures/gsea_down_plot_2.0.png"), gsea_down_plot, width = 6.5, height = 5)
```

Now the same for increasing pathways

```{r}
up_pathways_to_show <- data.frame(full_name = c("GOBP_CELLULAR_RESPIRATION",
                         "GOBP_PROTON_MOTIVE_FORCE_DRIVEN_ATP_SYNTHESIS",
                         "GOBP_ELECTRON_TRANSPORT_CHAIN",
                         "GOCC_RESPIRATORY_CHAIN_COMPLEX",
                         "HALLMARK_OXIDATIVE_PHOSPHORYLATION",
                         "GOBP_OXIDATIVE_PHOSPHORYLATION"),
                         plot_name = c("Cellular respiration (GOBP)",
                                       "Proton motive force driven ATP synthesis (GOBP)",
                                       "Electron transport chain (GOBP)",
                                       "Respiratory chain complex (GOCC)",
                                       "Oxidative phosphorylation (Hallmark)",
                                       "Oxidative phosphorylation (GOBP)"))

gsea_up_df <- up_pathways_to_show %>%
  inner_join(
    gsea$UP@result %>% select(ID, p.adjust),
    by = c("full_name" = "ID")
  ) %>%
  filter(p.adjust < 0.05)

gsea_up_plot <- ggplot(data = gsea_up_df, 
                         aes(x = -log10(p.adjust),
                             y = reorder(plot_name, -p.adjust))) +
  geom_col(fill = "magenta3") +
  theme_bw() +
  ylab("Gene set") +
  xlab(expression("-log"[10]*"p-value")) +
  ggtitle("Gene sets with increased activity \nin jSSc monocytes over time since ASCT") +
  theme(
    text = element_text(face = "bold"),
    axis.text = element_text(color = "black", size = 13),
    axis.title = element_text(color = "black", size = 15),
    plot.title = element_text(color = "black", size = 15)
  ) +
  scale_y_discrete(labels = function(x) str_wrap(x, width = 20))

ggsave(paste0(project_path, "figures/gsea_up_plot_2.0.png"), gsea_up_plot, width = 6.5, height = 5)

```


Now, tile plot

```{r}
# Get all gene sets
gene_sets <- gsea$DOWN@geneSets
# Extract 5 favorite pathways
fav_gene_sets <- gene_sets[favorite_pathways]

# Make a named vector of p-values for quick lookup
pval_lookup <- results$pval
names(pval_lookup) <- results$name

# Order each pathway's genes by p-value
fav_gene_sets_ordered <- lapply(fav_gene_sets, function(genes) {
  # Keep only genes that exist in results
  genes_in_results <- intersect(genes, names(pval_lookup))
  
  # Order by p-value
  genes_in_results[order(pval_lookup[genes_in_results])]
})

topgenes <- lapply(fav_gene_sets_ordered, function(x){head(x, 5)}) %>% unlist() %>% setNames(NULL) %>% unique()

pathway_by_gene_mat <- sapply(topgenes, function(gene) {
  sapply(fav_gene_sets_ordered, function(x) gene %in% x)
})

heat_df <- pathway_by_gene_mat %>%
  as.data.frame() %>%
  rownames_to_column("pathway") %>%
  pivot_longer(-pathway, names_to = "gene", values_to = "present")

heat_df$gene <- factor(heat_df$gene, levels = topgenes)

heat_df <-
heat_df %>%
    left_join(
        gsea_down_df %>% select(full_name, plot_name),
        by = c("pathway" = "full_name")
    )


gsea_plot_with_genes <- 
  ggplot(heat_df, aes(x = gene, y = plot_name, fill = present)) +
    geom_tile(color = "white") +
    scale_fill_manual(values = c("FALSE" = "white", "TRUE" = "firebrick")) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid = element_blank(),
      legend.position = "none",
      text = element_text(face = "bold"),
    axis.text = element_text(color = "black", size = 13),
    axis.title = element_text(color = "black", size = 15),
    plot.title = element_text(color = "black", size = 15)
    ) +
    ylab("Gene Set") +
    xlab("Gene") +
    ggtitle("Top DEGs contributing to pathways with decreased \nactivity in jSSc monocytes over time since ASCT")+
  scale_y_discrete(labels = function(x) str_wrap(x, width = 25))

ggsave(paste0(project_path, "figures/gsea_down_plot_with_genes_2.0.png"), gsea_plot_with_genes, width = 8, height = 6, bg = "white")
```

Now, our longitudinal plots

```{r}
# Add LYN to SRF genes to have even six
# In SSc, the high expression of LYN in monocytes promotes their differentiation into fibrocytes.

srf_genes_2 <- c("SERPINE1", "FOSL1", "SMAD3", "TAGLN2", "CDKN1C", "LYN")

gene_plots <- plot_pb_offset(pbo_monocytes, srf_genes_2, groupBy = "subject")

gene_plots_clean <- lapply(seq_along(gene_plots), function(i) {
  p <- gene_plots[[i]]
  nrow <- 6; ncol <- 3
  row <- (i - 1) %/% ncol
  col <- (i - 1) %% ncol
  
  p <- p + theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    #plot.margin = margin(0.25, 0.25, 0.25, 0.25, "pt"),
    plot.title = element_text(size = 30)
  )
  
  if (row < nrow - 1) {
    #p <- p + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
  }
  
  p
})

gene_plots_combined <- wrap_plots(gene_plots_clean, nrow = 2, ncol = 3) +
  plot_layout(guides = "collect") +
  plot_annotation(
    caption = "Months since transplant"
  ) &
  theme(
    legend.position = "right",
    legend.text = element_text(size = 30, face = "plain"),   # fix bold/oversized
    legend.title = element_text(size = 30, face = "plain"),
    legend.key.size = unit(0.8, "lines"),
    plot.caption = element_text(hjust = 0.5, size = 30),
    plot.margin = margin(50, 10, 10, 30, "pt") 
  )

gene_plots_combined <- wrap_elements(full = gene_plots_combined) +
  labs(tag = "Counts per million (log2 scale)") +
  theme(
    plot.tag = element_text(size = 30, angle = 90, hjust = 0.5, vjust = -1),
    plot.tag.position = "left"
  ) + ggtitle("Downward longitudinal trends for key DEGs in jSSc monocytes over \ntime since ASCT") + theme(plot.title = element_text(size = 30), face = "bold")

ggsave(paste0(project_path, "figures/gene_plots_combined.png"), 
       gene_plots_combined, width = 14, height = 8, bg = "white")

```

